### 2025-04-15 about.qmd
Initial commit

---
title: "Members"
---

[Marianne Huebner](https://cstat.msu.edu/people/marianne-huebner-phd-sheher) (chair), Professor of Statistics and Probability at Michigan State University, USA. <https://orcid.org/0000-0002-9694-9231>

Carsten Oliver Schmidt (chair), Professor of the Institute of Community Medicine, University Medicine of Greifswald

Mark Baillie, Advanced Methodology and Data Science, Clinical Development and Analytics. Novartis Pharma AG, Basel, Switzerland

Saskia le Cessie, Professor at Leiden University Medical Center, Netherlands

[Lara Lusa](https://www.famnit.upr.si/sl/zaposleni-in-sodelavci/lara.lusa/), Professor at the Faculty of Mathematics, Natural Sciences and Information Technologies of the University of Primorska in Koper, Slovenia

[Michael Schell](https://www.moffitt.org/research-science/researchers/michael-schell/), Moffit Cancer Center, Florida, USA

## **Bios**

**Mark Baillie** is a statistician supporting the clinical development and analytics department at Novartis. He is a member of the methodology and consulting group, providing strategic and methodological input to various clinical and non-clinical priority projects across the company. He has a focus on effective visual communication working on a number of internal and external initiatives to improve the reporting of clinical trials and exploratory projects.

**Marianne Huebner** is Professor of the Department of Statistics and Probability and Director of the Center for Statistical Training and Consulting at Michigan State University. After receiving a PhD in Applied Mathematics she held positions at UC Berkeley, the Mathematical Sciences Research Institute in Berkeley, the Mayo Clinic, and the University Medical Center in Hamburg, Germany. She develops and applies statistical models motivated by scientific questions. Her research interests include modeling health outcomes, initial data analyses, and sports and exercise.

**Saskia le Cessie** is a statistician with a joint appointment at the Department of Clinical Epidemiology and the Section Medical Statistics, Department of Biomedical Datasciences. She obtained a master in Mathematics (with minor in Computer Sciences) at the University of Utrecht and obtained her PhD in Medical Statistics at the University of Leiden. She is a broadly oriented statistician and involved in several large epidemiological studies of the Department of Clinical Epidemiology. Her research interests are in epidemiological and statistical methods for observational studies, in particular instrumental variable analysis, mediation analysis, competing risks, causal modelling, meta analysis and repeated measurements.

**Lara Lusa** is Professor of Statistics at the Faculty of Mathematics, Natural Sciences and Information Technologies of the University of Primorska, Slovenia. After receiving a PhD in Applied Statistics from the University of Florence, she worked at the Italian National Cancer Institute of Milan, at the Institute of Biostatistics and Medical Informatics of the University of Ljubljana and was a visiting fellow at the Biometric Research Branch of the National Cancer Institute in Bethesda. Her current research interests are related to the development and validation of predictive models in bio‐medicine, with particular attention to applications using high‐dimensional data and rare events. She is also interested in statistical programming, design of simulation studies, and teaching statistics. Beside the methodological research in statistics, she is also heavily involved as a statistical consultant to researchers in medicine.

**Michael Schell**

**Carsten Oliver Schmidt** is Professor at the Institute of Community Medicine, University Medicine of Greifswald. He is deputy head of the Department of SHIP/ Clinical-Epidemiological Research and leads the functional division "Quality in the Health Sciences". He is PI of several methods and research quality related projects, and speaker of the methods section of the German Society for Epidemiology. He is responsible for quality management procedures in population based observational health studies, has expertise in data linkage of primary and secondary data including registries, claims data, hospitals, and in the development of research infrastructures. Research topics include research methods, the study of subclinical and clinical disorders, incidental findings.

### 2025-04-15 activities.qmd
Initial commit

---
title: "Activities"
---

## IDA glossary

### **Project description**

The objective is to provide definitions of common terms used by the Initial Data Analysis topics group. Every effort will be made to use existing opne access definitions.

Authors: All TG3 members

Project status: **NEW**

### **Project outcomes:**

Glossary of 10-15 terms with definitions, e.g IDA, EDA, data cleaningm data screening, data quality,

## Incorporating IDA in courses on statistics

### **Project description**

The objective is to provide an engaging application from itial data analysis to regression models using characters from the Pokémon franchise to address the research question aimed at evaluating the association between Pokémon species weight and other characteristics.

This example provides flexibility for instructors (courses, workshops, seminars, modules) to integrate into their teaching materials or assignments.  Depending on the type of classes, the instructor can emphasize different aspects, for example data descriptions, learning which explorations are meaningful and how to interpret the results, univariable or multivariable regression modeling strategies, learning coding in R, or rigor and reproducibility in planning and executing studies.

Lead: Lara Lusa

Project status: **ACTIVE**

### **Project outcomes:**

1.\ Manuscript for a journal with a "Teaching corner"

2.\ Curated Pokemon data set for open access

3.\ R code for initial data analysis and for regression models

4.\ Statistical analysis plan for this example

5.\ Diskussion questions or tasks for the students

## IDA for regression models

### **Project description**

The objective is to provide examples on conducting initial data analysis (IDA) in a reproducible manner in the context of intended regression analyses.

Authors: Georg Heinze, Mark Baillie, Lara Lusa, Carsten O Schmidt, Willi Sauerbrei, Frank E Harrell, Marianne Huebner

This is a joint project between the Topic Groups 2 and 3 of the STRATOS Initiative.

Project status: **CLOSED, paper published**

### **Project outcomes:**

1.\  Open access dataset: [bacteremia](https://zenodo.org/records/7554815)

2.\  [IDA report with R code](https://stratosida.github.io/regression-regrets/)

3.\  IDA plan (check list) for the example(s) (in the paper)

4.\  [Paper](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-024-02294-3)  with scope of regression model, results, and consequences of IDA for regression:

Heinze G, Baillie M, Lusa L, Sauerbrei W, Schmidt CO, Harrell FE, Huebner M; TG2 and TG3 of the STRATOS initiative. Regression without regrets -initial data analysis is a prerequisite for multivariable regression. BMC Med Res Methodol. 2024 Aug 8;24(1):178. [doi: 10.1186/s12874-024-02294-3](https://doi.org/10.1186/s12874-024-02294-3)

## IDA for longitudinal data

### **Project description**

The objective is to provide an example of initial data anlysis for longitudinal data.

Authors: Lara Lusa, Cecile Proust-Lima, Katherine Lee, Frank Lawrence, Carsten Schmidt, Mark Baillie, Saskia le Cessie, Marianne Huebner

Project status: **CLOSED, paper published**

### **Project outcomes:**

1.\  IDA plan [check list](assets/IDAchecklist_long.pdf)

2.\  [IDA report with R code](https://stratosida.github.io/longitudinal/)

3.\  Manuscript with IDA workflow in the example of the SHARE data:

Lusa L, Proust-Lima C, Schmidt CO, Lee KJ, le Cessie S, Baillie M, Lawrence F, Huebner M; TG3 of the STRATOS Initiative. Initial data analysis for longitudinal studies to build a solid foundation for reproducible analysis. PLoS One. 2024 May 29;19(5):e0295726. [doi: 10.1371/journal.pone.0295726](https://doi.org/10.1371/journal.pone.0295726)

### **Supplementary material**

The Survey of Health Ageing and Retirement in Europe (SHARE) consists of data on health and socioeconomic variables of non-institutionalized individuals aged 50 and older across several European countries. The data sets includes about 140 000 men and women, ages 50 or older, collected in years 2004 to 2018. The questionnaires include questions on demographics, health (physical, and mental health, cognitive function, grip strength and behavioral risks), and socio-economic status. <https://share-eric.eu>

**Open Science Framework: R code**

Lusa L, Huebner M. (2021, September 17). Repository with R vignettes to support data organization and analysis for the SHARE study. <https://doi.org/10.17605/OSF.IO/KGTX6>

## Systematic review of IDA reporting

### **Project description**

Our aim was to describe the practice of reporting of initial data analyses in observational studies in five highly ranked medical journals with focus on data cleaning, screening, and reporting of findings which led to a potential change in the analysis plan.

Authors: Marianne Huebner, Werner Vach, Saskia le Cessie, Carsten Oliver Schmidt, Lara Lusa

Project status: **CLOSED, paper published**

### **Project outcomes**

1\. Summary of current IDA reporting by location in papers with examples.

2\. Recommendations for reporting practice of IDA in papers

3\. Manuscript:

Hidden analyses: a review of reporting practice and recommendations for more transparent reporting of initial data analyses

BMC Med Res Meth (2020) 20:61. <https://doi.org/10.1186/s12874-020-00942-y>

### **Supplementary material**

In the [assets](https://github.com/stratosida/stratosida.github.io/tree/main/assets/) folder

1\. Study protocol for IDA literature review: [Protocol_IDA20180620.pdf](assets/Protocol_IDA20180620.pdf)

2\. Design of IDA collection form: [Collection_Form_20181129.pdf](assets/Collection_Form_20181129.pdf)

### 2025-04-15 glossary.qmd
Initial commit

---
title: "Glossary"
---

## Initial data analysis framework

### Initial Data Analysis

The main aim of IDA is seen in providing reliable knowledge about the data to enable responsible statistical analyses and interpretation. 
IDA has the following phases: (1) metadata setup; (2) data cleaning; (3) data screening; (4) initial data reporting; 
(5) refining and updating the research analysis plan; and (6) documenting and reporting IDA. IDA is aligned with the research aims 
and the statistical analysis plan and does not include hypothesis generating activities.

**Source:** Schmidt CO, Vach W, le Cessie S, Huebner M. STRATOS: Introducing the Initial Data Analysis Topic Group (TG3). Biometric Bulletin 2018; 35 (2): 10-11

[Link](https://www.stratos-initiative.org/sites/default/files/2021-11/STRATOS-TG3-BB-initial-data-analysis.pdf)

### Exploratory Data Analysis

Exploratory Data Analysis (EDA) inspects data to uncover patterns in the data or identifies possible errors and anomalies, for example through graphical approaches. EDA may include investigations for model building and hypothesis generation in a cyclical manner.

**Source:** Lusa L, Proust-Lima C, Schmidt CO, Lee KJ, le Cessie S, Baillie M, Lawrence F, Huebner M; TG3 of the STRATOS Initiative. Initial data analysis for longitudinal studies to build a solid foundation for reproducible analysis. PLoS One. 2024 May 29;19(5):e0295726. doi: 10.1371/journal.pone.0295726.

[Link](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0295726)

**References:**

Behrens JT. Principles and procedures of exploratory data analysis. Psychological methods. 1997;2(2):131. <https://core.ac.uk/download/pdf/193648223.pdf>

Cook D, Swayne DF. Interactive and Dynamic Graphics for Data Analysis: With R and Ggobi. New York, NY: Springer New York; 2007. <https://en.wikipedia.org/wiki/Exploratory_data_analysis>

### Statistical Analysis Plan (SAP)

A statistical analysis plan is a document that contains a more technical and detailed elaboration of the principal features of the analysis described in the protocol and includes detailed procedures for executing the statistical analysis of the primary and secondary variables and other data.

**Source:** International Council for Harmonisation (ICH E9). ICH: E 9: Statistical principles for clinical trials - Step 5. Glossary p. 37

[Link](https://www.ema.europa.eu/en/ich-e9-statistical-principles-clinical-trials-scientific-guideline)

### Initial Data Analysis Plan (IDAP)

An IDA plan is focused on data screening to prepare for statistical analyses; it does not specify procedures or software packages to conduct the analyses. It can be
understood as a minimum basic set of analyses that can be extended depending on the context of the research study and data collection. IDA domains missing values, univariable and multivariable descriptions.

**Source:** [Heinze et al https://doi.org/10.1186/s12874-024-02294-3;  Lusa et al https://doi.org/10.1371/journal.pone.0295726

## Elements of initial data analysis

### Data wrangling

Data Wrangling is a broad term referring to the processes involved when preparing data for analysis. It can include acquiring data, enriching, changing the format and shape of the data, combining, subsetting and sampling data, and cleaning data. Some common steps involved with Data Wrangling are:

* Discovering and gathering the data needed 
* Merging data from different sources, if necessary  
* Fixing flaws in the data entries  
* Extracting the necessary data and put it in the proper structure  
* Storing it in the proper format for further use

**Source:** National Library of Medicine

[Link](https://www.nnlm.gov/guides/data-glossary/data-wrangling)

### Data cleaning

Data cleaning is the process of identifying and correcting data that are inaccurate, missing, or incomplete. Data cleaning tasks can include removing duplicate records, investigating extreme values (e.g., outliers), converting dates from one format to another, removing unwanted text, splitting multiple data points in a cell into separate cells, or coding missing or NA values.

**Source:** National Library of Medicine

[Link](https://www.nnlm.gov/guides/data-glossary/data-cleaning)

### Data screening

Data screening consists of reviewing and documenting the properties and quality of the data that may affect future analysis and interpretation.

**Source:** Huebner M, le Cessie S, Schmidt CO, Vach W. A contemporary conceptual framework for initial data analysis. Observational Studies 2018; 4: 171-192.

[Link](https://doi.org/10.1353/obs.2018.0014)

### Structural variables

Structural variables are variables that help to structure initial data analysis results for a clear organization and essential overview of data properties. Structuring can be based on levels of measurement (centers), on calendar time of recruitment, on demographic variables such as sex or age, or on variables of central importance to the research questions. Structural variables may or may not be included as predictors in the statistical models.

**Source:** [Heinze et al https://doi.org/10.1186/s12874-024-02294-3]

## Data and related terms

### Data

Reinterpretable representation of information in a formalized manner suitable for communication, interpretation, or processing

**Source:** ISO/IEC 2382:2015, 2121272, modified

[Link](https://www.iso.org/obp/ui/#iso:std:iso:8000:-2:ed-2:v1:en)

### Data set

logically meaningful grouping of data

 **Source:**

[Link](https://www.iso.org/obp/ui/#iso:std:iso:8000:-2:ed-2:v1:en)

### Metadata

data defining and describing other data

**Source:** ISO/IEC 11179‑1:2015, 3.2.16

[Link](https://www.iso.org/obp/ui/#iso:std:iso:8000:-2:ed-2:v1:en)

### Data dictionary

A data dictionary is a document that outlines the structure, content, and meaning of a given variable. This includes what type of data is being collected (e.g. free text, numerical, categorical or group data), the full wording of a question, what values are allowable (e.g. numeric ranges, multiple choice codes), and what those values mean (e.g. 0 = no high blood pressure diagnosis, 1 = borderline high blood pressure, 2 = high blood pressure). A data dictionary is a critical tool for data analysis and reproducibility.

**Source:** National Library of Medicine

[Link-NLM](https://www.nnlm.gov/guides/data-glossary/data-dictionary)

[Link-USGS](https://www.usgs.gov/data-management/data-dictionaries)

### Data quality

Data quality is the degree to which a set of characteristics of data fulfils requirements (=need or expectation that is stated, generally implied or obligatory)

**Source:** ISO 9000:2015, 3.6.2  and  ISO 9000:2015, 3.6.4

[Link1](https://www.iso.org/obp/ui/#iso:std:iso:8000:-2:ed-2:v1:en)

[Link2](https://www.iso.org/obp/ui/#iso:std:iso:8000:-2:ed-2:v1:en)

### Data subject

A Data subject is the individual the personal data relates to.

**Source:** General Data Protection Regulation (GDPR)

[Link](https://www.edpb.europa.eu/system/files/2023-06/Key%20GDPR%20Definitions%20Infographic.pdf)

### Data provenance

The term “data provenance”, sometimes called “data lineage,” refers to a documented trail that accounts for the origin of a piece of data and where it has moved from to where it is presently. The purpose of data provenance is to tell researchers the origin, changes to, and details supporting the confidence or validity of research data. The concept of provenance guarantees that data creators are transparent about their work and where it came from and provides a chain of information where data can be tracked as researchers use other researchers’ data and adapt it for their own purposes.

**Source:** National Library of Medicine

[Link](https://www.nnlm.gov/guides/data-glossary/data-provenance)

### 2025-04-15 index.qmd
Initial commit

---
title: "Initial Data Analysis"
---

## Aims

To improve awareness of Initial Data Analysis (IDA) as an important part of the research process and to provide guidance on conducting IDA in a systematic and reproducible manner.

## Key principle

IDA does not look at associations of covariates with the outcome. The aim is to provide reliable knowledge about the data to ensure transparency and integrity of preconditions to conduct appropriate statistical analyses and correct interpretation of the results to answer pre-defined research questions.

## Definition and IDA Framework

Initial Data Analysis (IDA) consists of all steps performed on the data of a study typically between the end of the data collection/entry and start of those statistical analyses that address research questions. 
A systematic plan of IDA activities (IDAP) should be incoporated in the statistical analysis plan (SAP). Ideally, IDA should already be performed during ongoing data collections to detect and deal with data issues as early as possible.

![](assets/idaframework.png)

## STRATOS Initiative

Topic group 3 is a member of the [STRATOS](https://stratos-initiative.org) Initiative (STRengthening Analytical Thinking for Observational Studies) which is a large collaboration of experts in many different areas of biostatistical research. Ongoing research, discussions and activities within STRATOS are conducted in nine [topic groups](https://stratos-initiative.org/en/groups) and several cross-cutting [panels](https://stratos-initiative.org/en/panels).

![Conference BIRS 2016](assets/Stratos_group2016.jpg){fig-align="center"}

### 2025-04-15 members.qmd
Initial commit

--- 2025-04-15 about.qmd
@@
 
-[Marianne Huebner](https://cstat.msu.edu/people/marianne-huebner-phd-sheher) (chair), Professor of Statistics and Probability at Michigan State University, USA. <https://orcid.org/0000-0002-9694-9231>
+[Marianne Huebner](https://cstat.msu.edu/people/marianne-huebner-phd-sheher) (chair), Professor of Statistics and Probability at Michigan State University, USA.
+
+[Lara Lusa](https://www.famnit.upr.si/sl/zaposleni-in-sodelavci/lara.lusa/),(chair) Professor at the Faculty of Mathematics, Natural Sciences and Information Technologies of the University of Primorska in Koper, Slovenia
 
@@
 Saskia le Cessie, Professor at Leiden University Medical Center, Netherlands
-
-[Lara Lusa](https://www.famnit.upr.si/sl/zaposleni-in-sodelavci/lara.lusa/), Professor at the Faculty of Mathematics, Natural Sciences and Information Technologies of the University of Primorska in Koper, Slovenia
 
@@
 
-**Marianne Huebner** is Professor of the Department of Statistics and Probability and Director of the Center for Statistical Training and Consulting at Michigan State University.
+**Marianne Huebner** is a professor of the Department of Statistics and Probability and Director of the Center for Statistical Training and Consulting at Michigan State University.
 After receiving a PhD in Applied Mathematics she held positions at UC Berkeley, the Mathematical Sciences Research Institute in Berkeley, the Mayo Clinic, and the University Medical Center in Hamburg, Germany.
@@
 Her research interests include modeling health outcomes, initial data analyses, and sports and exercise.
+<https://orcid.org/0000-0002-9694-9231>
 
@@
 
-**Lara Lusa** is Professor of Statistics at the Faculty of Mathematics, Natural Sciences and Information Technologies of the University of Primorska, Slovenia.
+**Lara Lusa** is a professor of Statistics at the Faculty of Mathematics, Natural Sciences and Information Technologies of the University of Primorska, Slovenia.
 After receiving a PhD in Applied Statistics from the University of Florence, she worked at the Italian National Cancer Institute of Milan, at the Institute of Biostatistics and Medical Informatics of the University of Ljubljana and was a visiting fellow at the Biometric Research Branch of the National Cancer Institute in Bethesda.
@@
 Research topics include research methods, the study of subclinical and clinical disorders, incidental findings.
+<https://orcid.org/0000-0001-5266-9396>

### 2025-04-15 papers.qmd
Initial commit

---
title: "Publications"
---
## **TG3 Publications**

*1. This is a great overview over initial data analysis and the steps to think about .*

Baillie M, le Cessie S, Schmidt CO, Lusa L, Huebner M. Ten simple rules for initial data analysis. PLoS Comput Biol 2022; 18(2): e1009819. <https://doi.org/10.1371/journal.pcbi.1009819>

*2. This is the main paper that introduces the initial data analysis framework and explains concepts.*

Huebner M, le Cessie S, Schmidt CO, Vach W . A contemporary conceptual framework for initial data analysis. Observational Studies 2018; 4: 171-192. <https://doi.org/10.1353/obs.2018.0014>

*3. IDA check list for cross-sectional studies with example and code:*

Heinze G, Baillie M, Lusa L, Sauerbrei W, Schmidt CO, Harrell FE, Huebner M; TG2 and TG3 of the STRATOS initiative. Regression without regrets -initial data analysis is a prerequisite for multivariable regression. BMC Med Res Methodol. 2024 Aug 8;24(1):178. <https://doi.org/10.1186/s12874-024-02294-3>

IDA checklist for cross-sectional studies: [Link](assets/IDAchecklist_cross.pdf)

*4. IDA check list for longitudinal studies with example and code:*

Lusa L, Proust-Lima C, Schmidt CO, Lee KJ, le Cessie S, Baillie M, Lawrence F, Huebner M. Initial data analysis for longitudinal studies to build a solid foundation for reproducible analysis. PLoS One. 2024 May 29;19(5):e0295726. <https://doi.org/10.1371/journal.pone.0295726>

IDA checklist for longitudinal studies: [Link](https://stratosida.github.io/assets/IDAchecklist_long.pdf)

*5. This is a look at reporting practices and recommendations for more transparency.*

Huebner M, Vach W, le Cessie S, Schmidt C, Lusa L. Hidden Analyses: a review of reporting practice and recommendations for more transparent reporting of initial data analyses. BMC Med Res Meth 2020; 20:61. <https://doi.org/10.1186/s12874-020-00942-y>

*6. Our first attempt at explaining initial data analysis aimed at a clinical audience. This is superseded by the framework paper \[2\].*

Huebner M, Vach W, le Cessie S. A systematic approach to initial data analysis is good research practice. J Thoracic Cardiovas S. 2016; 151(1): 25-27. [Link](https://www.google.com/url?q=https%3A%2F%2Fwww.jtcvs.org%2Farticle%2FS0022-5223%2815%2901794-8%2Fpdf&sa=D&sntz=1&usg=AOvVaw1nfle1UGmUgJMRaITfCUEU)

*7. Short introductions of activities of our topic group :*

-   *Initial:* Schmidt CO, Vach W, le Cessie S, Huebner M. STRATOS: Introducing the Initial Data Analysis Topic Group (TG3). Biometric Bulletin 2018; 35 (2): 10-11.

-   *Update:* Le Cessie S, Schmidt CO, Lusa L, Baillie M, Huebner M. STRengthening Analytical Thinking for Observational Studies (STRATOS): Introducing the Initial Data Analysis Topic Group (TG3). Biometric Bulletin 2021; 38 (3): 10-11.

-   *Data quality:* Schmidt CO, Heinze G, Lusa L, Huebner M. STRengthening Analytical Thinking for Observational Studies (STRATOS): On the importance of data quality assessment and initial data analysis. Biometric Bulletin 2022; 39 (4): 14-15.

-   *Glossary:* Boeker M, Tippmann P, Day S, Huebner M, Sauerbrei W on behalf of the Glossary Panel: STRengthening Analytical Thinking for Observational Studies (STATOS): Defining common meaning for statistical Terms. Biometrical Bulletin; 37(4): 13-14.

## **Related Publications. Collaborations with TG3 members**

1.  *This paper provides an extensive overview of data quality aspects on observational studies and provides R code to conduct data quality assessments.*

Schmidt CO, Struckmann S, Enzenbach C, Reineke A, Stausberg J, Damerow S, Huebner M, Schmidt B, Sauerbrei W, Richter A. Facilitating harmonized data quality assessments. A data quality framework for observational health research data collections with software implementations in R. BMC Med Res Meth 2021; 21(1): 1-15. [Link](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-021-01252-7)

2.  *An overview of an R package to provide data quality assessments*

Richter A , Schmidt CO , Krüger M , and Struckmann S. dataquieR: assessment of data quality in epidemiological research. Journal of Open Source Software 2021; 6(61), 3093, <https://doi.org/10.21105/joss.03093>

3.  *Data management and data cleaning for a longitudinal survey design*

Lusa L and Huebner M. Organizing and Analyzing Data from the SHARE Study with an Application to Age and Sex Differences in Depressive Symptoms. Int. J. Environ. Res. Public Health 2021, 18(18), 9684; <https://doi.org/10.3390/ijerph18189684>

4.  *This is the accompanying R code for the IJEPRH paper (item 3)*

Lusa L and Huebner M. Repository with R vignettes to support data organization and analysis for the SHARE study. 2021. <https://doi.org/10.17605/OSF.IO/KGTX6>

### 2025-04-15 resources.qmd
Initial commit

---
title: "R implementations"
---

### IDA report for regression models

[Bacteremia data set](https://zenodo.org/records/7554815) by Georg Heinze

<https://stratosida.github.io/regression-regrets/>

### IDA report for longitudinal studies

Grip strength data set from the [SHARE project](https://share-eric.eu)

<https://stratosida.github.io/longitudinal/>

### 2025-04-16 index.qmd
Update index.qmd
Update index.qmd content
Update index.qmd with TG2 content

@@
 ---
-title: "Initial Data Analysis"
+title: "Selection of Variables and Functional Forms in Multivariable Analysis"
 ---
@@
 
-To improve awareness of Initial Data Analysis (IDA) as an important part of the research process and to provide guidance on conducting IDA in a systematic and reproducible manner.
+Derive guidance for variable and function selection  in multivariable analysis.
 
-## Key principle
+## Focus
 
-IDA does not look at associations of covariates with the outcome.
-The aim is to provide reliable knowledge about the data to ensure transparency and integrity of preconditions to conduct appropriate statistical analyses and correct interpretation of the results to answer pre-defined research questions.
-
-## Definition and IDA Framework
-
-Initial Data Analysis (IDA) consists of all steps performed on the data of a study typically between the end of the data collection/entry and start of those statistical analyses that address research questions.
-A systematic plan of IDA activities (IDAP) should be incoporated in the statistical analysis plan (SAP).
-Ideally, IDA should already be performed during ongoing data collections to detect and deal with data issues as early as possible.
-
-![](assets/idaframework.png)
+The main focus of TG2 is to identify influential variables and gain insight into their individual and joint relationship with the outcome.
+Two of the (interrelated) main challenges are selection of variables for inclusion in a multivariable explanatory model, and choice of the functional forms for continuous variables
 
@@
 
-Topic group 3 is a member of the [STRATOS](https://stratos-initiative.org) Initiative (STRengthening Analytical Thinking for Observational Studies) which is a large collaboration of experts in many different areas of biostatistical research.
+Topic group 2 is a member of the [STRATOS](https://stratos-initiative.org) Initiative (STRengthening Analytical Thinking for Observational Studies) which is a large collaboration of experts in many different areas of biostatistical research.
 Ongoing research, discussions and activities within STRATOS are conducted in nine [topic groups](https://stratos-initiative.org/en/groups) and several cross-cutting [panels](https://stratos-initiative.org/en/panels).
 
-![Conference BIRS 2016](assets/Stratos_group2016.jpg){fig-align="center"}
+![2nd General Meeting of the STRATOS members at the Banff International Research Station (BIRS), Banff, Canada.](assets/STRATOS_Banff_second_meeting.png){fig-align="center"}

### 2025-04-16 members.qmd
Update members.qmd with TG2 content

--- 2025-04-15 about.qmd
@@
 
-[Marianne Huebner](https://cstat.msu.edu/people/marianne-huebner-phd-sheher) (chair), Professor of Statistics and Probability at Michigan State University, USA. <https://orcid.org/0000-0002-9694-9231>
+[Georg Heinze](https://www.meduniwien.ac.at/researcher/georg_heinze) (chair): Institute of Clinical Biometrics, Center for Medical Statistics, Informatics and Intelligent Systems, Medical University of Vienna,  Austria.
 
-Carsten Oliver Schmidt (chair), Professor of the Institute of Community Medicine, University Medicine of Greifswald
+**Aris Perperoglou (chair)**:  Data Science & AI,   AstraZeneca R&D Biopharmaceuticals , Cambridge UK
 
-Mark Baillie, Advanced Methodology and Data Science, Clinical Development and Analytics. Novartis Pharma AG, Basel, Switzerland
+**Willi Sauerbrei (chair)**: Institute for Medical Biometry and Medical Informatics, Medical Center - University of Freiburg, Germany
 
-Saskia le Cessie, Professor at Leiden University Medical Center, Netherlands
+**Michal Abrahamowicz**: Faculty of Medicine, McGill University, Canada
 
-[Lara Lusa](https://www.famnit.upr.si/sl/zaposleni-in-sodelavci/lara.lusa/), Professor at the Faculty of Mathematics, Natural Sciences and Information Technologies of the University of Primorska in Koper, Slovenia
+**Heiko Becher**: Institute for Medical Biometry and Epidemiology, University Medical Center Hamburg-Eppendorf, Hamburg, Germany
 
-[Michael Schell](https://www.moffitt.org/research-science/researchers/michael-schell/), Moffit Cancer Center, Florida, USA
+**Harald Binder**:  Institute for Medical Biometry and Medical Informatics, Medical Center - University of Freiburg, Germany
 
-## **Bios**
+**Daniela Dunkler**: Institute of Clinical Biometrics, Center for Medical Statistics, Informatics and Intelligent Systems, Medical University of Vienna,  Austria
 
-**Mark Baillie** is a statistician supporting the clinical development and analytics department at Novartis.
-He is a member of the methodology and consulting group, providing strategic and methodological input to various clinical and non-clinical priority projects across the company.
-He has a focus on effective visual communication working on a number of internal and external initiatives to improve the reporting of clinical trials and exploratory projects.
+**Rolf Groenwold**:  Leiden University Medical Centre, Leiden, The Netherlands
 
-**Marianne Huebner** is Professor of the Department of Statistics and Probability and Director of the Center for Statistical Training and Consulting at Michigan State University.
-After receiving a PhD in Applied Mathematics she held positions at UC Berkeley, the Mathematical Sciences Research Institute in Berkeley, the Mayo Clinic, and the University Medical Center in Hamburg, Germany.
-She develops and applies statistical models motivated by scientific questions.
-Her research interests include modeling health outcomes, initial data analyses, and sports and exercise.
+**Frank Harrell**: School of Medicine, Vanderbilt University Medical Center, USA
 
-**Saskia le Cessie** is a statistician with a joint appointment at the Department of Clinical Epidemiology and the Section Medical Statistics, Department of Biomedical Datasciences.
-She obtained a master in Mathematics (with minor in Computer Sciences) at the University of Utrecht and obtained her PhD in Medical Statistics at the University of Leiden.
-She is a broadly oriented statistician and involved in several large epidemiological studies of the Department of Clinical Epidemiology.
-Her research interests are in epidemiological and statistical methods for observational studies, in particular instrumental variable analysis, mediation analysis, competing risks, causal modelling, meta analysis and repeated measurements.
+**Nadja Klein**: School of Business and Economics , Humboldt-University of Berlin, Germany
 
-**Lara Lusa** is Professor of Statistics at the Faculty of Mathematics, Natural Sciences and Information Technologies of the University of Primorska, Slovenia.
-After receiving a PhD in Applied Statistics from the University of Florence, she worked at the Italian National Cancer Institute of Milan, at the Institute of Biostatistics and Medical Informatics of the University of Ljubljana and was a visiting fellow at the Biometric Research Branch of the National Cancer Institute in Bethesda.
-Her current research interests are related to the development and validation of predictive models in bio‐medicine, with particular attention to applications using high‐dimensional data and rare events.
-She is also interested in statistical programming, design of simulation studies, and teaching statistics.
-Beside the methodological research in statistics, she is also heavily involved as a statistical consultant to researchers in medicine.
+**Geraldine Rauch**:  Institute of Biometry and Clinical  Epidemiology, Charité – Universitätsmedizin Berlin , Germany
 
-**Michael Schell**
+**Patrick Royston**: Department of Statistical Science, UCL, UK
 
-**Carsten Oliver Schmidt** is Professor at the Institute of Community Medicine, University Medicine of Greifswald.
-He is deputy head of the Department of SHIP/ Clinical-Epidemiological Research and leads the functional division "Quality in the Health Sciences".
-He is PI of several methods and research quality related projects, and speaker of the methods section of the German Society for Epidemiology.
-He is responsible for quality management procedures in population based observational health studies, has expertise in data linkage of primary and secondary data including registries, claims data, hospitals, and in the development of research infrastructures.
-Research topics include research methods, the study of subclinical and clinical disorders, incidental findings.
+**Matthias Schmid**:  Institute for Medicine Biometry, Informatics and Epidemiology, Bonn, Germany
+
+## Early career adjunct members
+
+**Michael Kammer**: Institute of Clinical Biometrics, Center for Medical Statistics, Informatics and Intelligent Systems, Medical University of Vienna,  Austria
+
+**Edwin  Kipruto**:  Institute for Medical Biometry and Medical Informatics, Medical Center - University of Freiburg, Germany
+
+**Kim Luijken**:  Leiden University Medical Centre, Leiden, The Netherlands
+
+**Christine Wallisch**: Institute of Clinical Biometrics, Center for Medical Statistics, Informatics and Intelligent Systems, Medical University of Vienna,  Austria

### 2025-04-16 members.qmd
Update members.qmd with TG2 content

@@
 
-[Georg Heinze](https://www.meduniwien.ac.at/researcher/georg_heinze) (chair): Institute of Clinical Biometrics, Center for Medical Statistics, Informatics and Intelligent Systems, Medical University of Vienna,  Austria.
+**[Georg Heinze](https://www.meduniwien.ac.at/researcher/georg_heinze) (chair)**: Institute of Clinical Biometrics, Center for Medical Statistics, Informatics and Intelligent Systems, Medical University of Vienna,  Austria.
 

### 2025-04-16 activities.qmd
Update activities.qmd with TG2 content

@@
 
-## IDA glossary
+NOTE: This is the text that was originally on the TG2 website and it is severely outdated!
 
-### **Project description**
+* Georg Heinze and Christine Wallisch will give a STRATOS-TG2 workshop on 'Variable selection – a review and recommendations for the practicing statistician' at the IBS-ROeS conference in Salzburg, 6 September 2021
 
-The objective is to provide definitions of common terms used by the Initial Data Analysis topics group. Every effort will be made to use existing opne access definitions.
+* Aris Perperoglou will host a session on "STRATOS initiative - on various issues in multivariable model building" at the RSS International Conference in Manchester,  6-9 September, with talks from Bianca de Stavola, Tim Morris, Willi Sauerbrei and Michal Abrahamowicz.
 
-Authors: All TG3 members
-
-Project status: **NEW**
-
-### **Project outcomes:**
-
-Glossary of 10-15 terms with definitions, e.g IDA, EDA, data cleaningm data screening, data quality,
-
-## Incorporating IDA in courses on statistics
-
-### **Project description**
-
-The objective is to provide an engaging application from itial data analysis to regression models using characters from the Pokémon franchise to address the research question aimed at evaluating the association between Pokémon species weight and other characteristics.
-
-This example provides flexibility for instructors (courses, workshops, seminars, modules) to integrate into their teaching materials or assignments.
-Depending on the type of classes, the instructor can emphasize different aspects, for example data descriptions, learning which explorations are meaningful and how to interpret the results, univariable or multivariable regression modeling strategies, learning coding in R, or rigor and reproducibility in planning and executing studies.
-
-Lead: Lara Lusa
-
-Project status: **ACTIVE**
-
-### **Project outcomes:**
-
-1.\ Manuscript for a journal with a "Teaching corner"
-
-2.\ Curated Pokemon data set for open access
-
-3.\ R code for initial data analysis and for regression models
-
-4.\ Statistical analysis plan for this example
-
-5.\ Diskussion questions or tasks for the students
-
-## IDA for regression models
-
-### **Project description**
-
-The objective is to provide examples on conducting initial data analysis (IDA) in a reproducible manner in the context of intended regression analyses.
-
-Authors: Georg Heinze, Mark Baillie, Lara Lusa, Carsten O Schmidt, Willi Sauerbrei, Frank E Harrell, Marianne Huebner
-
-This is a joint project between the Topic Groups 2 and 3 of the STRATOS Initiative.
-
-Project status: **CLOSED, paper published**
-
-### **Project outcomes:**
-
-1.\  Open access dataset: [bacteremia](https://zenodo.org/records/7554815)
-
-2.\  [IDA report with R code](https://stratosida.github.io/regression-regrets/)
-
-3.\  IDA plan (check list) for the example(s) (in the paper)
-
-4.\  [Paper](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-024-02294-3)  with scope of regression model, results, and consequences of IDA for regression:
-
-Heinze G, Baillie M, Lusa L, Sauerbrei W, Schmidt CO, Harrell FE, Huebner M; TG2 and TG3 of the STRATOS initiative.
-Regression without regrets -initial data analysis is a prerequisite for multivariable regression.
-BMC Med Res Methodol.
-2024 Aug 8;24(1):178.
-[doi: 10.1186/s12874-024-02294-3](https://doi.org/10.1186/s12874-024-02294-3)
-
-## IDA for longitudinal data
-
-### **Project description**
-
-The objective is to provide an example of initial data anlysis for longitudinal data.
-
-Authors: Lara Lusa, Cecile Proust-Lima, Katherine Lee, Frank Lawrence, Carsten Schmidt, Mark Baillie, Saskia le Cessie, Marianne Huebner
-
-Project status: **CLOSED, paper published**
-
-### **Project outcomes:**
-
-1.\  IDA plan [check list](assets/IDAchecklist_long.pdf)
-
-2.\  [IDA report with R code](https://stratosida.github.io/longitudinal/)
-
-3.\  Manuscript with IDA workflow in the example of the SHARE data:
-
-Lusa L, Proust-Lima C, Schmidt CO, Lee KJ, le Cessie S, Baillie M, Lawrence F, Huebner M; TG3 of the STRATOS Initiative.
-Initial data analysis for longitudinal studies to build a solid foundation for reproducible analysis.
-PLoS One.
-2024 May 29;19(5):e0295726.
-[doi: 10.1371/journal.pone.0295726](https://doi.org/10.1371/journal.pone.0295726)
-
-### **Supplementary material**
-
-The Survey of Health Ageing and Retirement in Europe (SHARE) consists of data on health and socioeconomic variables of non-institutionalized individuals aged 50 and older across several European countries.
-The data sets includes about 140 000 men and women, ages 50 or older, collected in years 2004 to 2018.
-The questionnaires include questions on demographics, health (physical, and mental health, cognitive function, grip strength and behavioral risks), and socio-economic status.
-<https://share-eric.eu>
-
-**Open Science Framework: R code**
-
-Lusa L, Huebner M. (2021, September 17). Repository with R vignettes to support data organization and analysis for the SHARE study. <https://doi.org/10.17605/OSF.IO/KGTX6>
-
-## Systematic review of IDA reporting
-
-### **Project description**
-
-Our aim was to describe the practice of reporting of initial data analyses in observational studies in five highly ranked medical journals with focus on data cleaning, screening, and reporting of findings which led to a potential change in the analysis plan.
-
-Authors: Marianne Huebner, Werner Vach, Saskia le Cessie, Carsten Oliver Schmidt, Lara Lusa
-
-Project status: **CLOSED, paper published**
-
-### **Project outcomes**
-
-1\. Summary of current IDA reporting by location in papers with examples.
-
-2\. Recommendations for reporting practice of IDA in papers
-
-3\. Manuscript:
-
-Hidden analyses: a review of reporting practice and recommendations for more transparent reporting of initial data analyses
-
-BMC Med Res Meth (2020) 20:61. <https://doi.org/10.1186/s12874-020-00942-y>
-
-### **Supplementary material**
-
-In the [assets](https://github.com/stratosida/stratosida.github.io/tree/main/assets/) folder
-
-1\. Study protocol for IDA literature review: [Protocol_IDA20180620.pdf](assets/Protocol_IDA20180620.pdf)
-
-2\. Design of IDA collection form: [Collection_Form_20181129.pdf](assets/Collection_Form_20181129.pdf)
+* Georg Heinze and Mark Baillie will host a session on 'STRATOS initiative - Tools for guidance for analysis of observational studies' at the IBS-ROeS conference in Salzburg, 7-10 September 2021, featuring Rolf Groenwold from TG2 with a talk on ' Using videos and other media for guidance'

### 2025-04-16 papers.qmd
Test papers.qmd collapsible update

@@
 ---
-title: "Publications"
+title: "TG2 Publications"
 ---
-## **TG3 Publications**
 
-*1. This is a great overview over initial data analysis and the steps to think about .*
-
-Baillie M, le Cessie S, Schmidt CO, Lusa L, Huebner M. Ten simple rules for initial data analysis. PLoS Comput Biol 2022; 18(2): e1009819. <https://doi.org/10.1371/journal.pcbi.1009819>
-
-*2. This is the main paper that introduces the initial data analysis framework and explains concepts.*
-
-Huebner M, le Cessie S, Schmidt CO, Vach W . A contemporary conceptual framework for initial data analysis. Observational Studies 2018; 4: 171-192. <https://doi.org/10.1353/obs.2018.0014>
-
-*3. IDA check list for cross-sectional studies with example and code:*
-
-Heinze G, Baillie M, Lusa L, Sauerbrei W, Schmidt CO, Harrell FE, Huebner M; TG2 and TG3 of the STRATOS initiative.
-Regression without regrets -initial data analysis is a prerequisite for multivariable regression.
-BMC Med Res Methodol.
-2024 Aug 8;24(1):178.
-<https://doi.org/10.1186/s12874-024-02294-3>
-
-IDA checklist for cross-sectional studies: [Link](assets/IDAchecklist_cross.pdf)
-
-*4. IDA check list for longitudinal studies with example and code:*
-
-Lusa L, Proust-Lima C, Schmidt CO, Lee KJ, le Cessie S, Baillie M, Lawrence F, Huebner M.
-Initial data analysis for longitudinal studies to build a solid foundation for reproducible analysis.
-PLoS One.
-2024 May 29;19(5):e0295726.
-<https://doi.org/10.1371/journal.pone.0295726>
-
-IDA checklist for longitudinal studies: [Link](https://stratosida.github.io/assets/IDAchecklist_long.pdf)
-
-*5. This is a look at reporting practices and recommendations for more transparency.*
-
-Huebner M, Vach W, le Cessie S, Schmidt C, Lusa L.
-Hidden Analyses: a review of reporting practice and recommendations for more transparent reporting of initial data analyses.
-BMC Med Res Meth 2020; 20:61.
-<https://doi.org/10.1186/s12874-020-00942-y>
-
-*6. Our first attempt at explaining initial data analysis aimed at a clinical audience. This is superseded by the framework paper \[2\].*
-
-Huebner M, Vach W, le Cessie S.
-A systematic approach to initial data analysis is good research practice.
-J Thoracic Cardiovas S.
-2016; 151(1): 25-27.
-[Link](https://www.google.com/url?q=https%3A%2F%2Fwww.jtcvs.org%2Farticle%2FS0022-5223%2815%2901794-8%2Fpdf&sa=D&sntz=1&usg=AOvVaw1nfle1UGmUgJMRaITfCUEU)
-
-*7. Short introductions of activities of our topic group :*
-
--   *Initial:* Schmidt CO, Vach W, le Cessie S, Huebner M. STRATOS: Introducing the Initial Data Analysis Topic Group (TG3). Biometric Bulletin 2018; 35 (2): 10-11.
-
--   *Update:* Le Cessie S, Schmidt CO, Lusa L, Baillie M, Huebner M.
-STRengthening Analytical Thinking for Observational Studies (STRATOS): Introducing the Initial Data Analysis Topic Group (TG3).
-Biometric Bulletin 2021; 38 (3): 10-11.
-
--   *Data quality:* Schmidt CO, Heinze G, Lusa L, Huebner M.
-STRengthening Analytical Thinking for Observational Studies (STRATOS): On the importance of data quality assessment and initial data analysis.
-Biometric Bulletin 2022; 39 (4): 14-15.
-
--   *Glossary:* Boeker M, Tippmann P, Day S, Huebner M, Sauerbrei W on behalf of the Glossary Panel: STRengthening Analytical Thinking for Observational Studies (STATOS): Defining common meaning for statistical Terms.
-Biometrical Bulletin; 37(4): 13-14.
-
-## **Related Publications. Collaborations with TG3 members**
-
-1.  *This paper provides an extensive overview of data quality aspects on observational studies and provides R code to conduct data quality assessments.*
-
-Schmidt CO, Struckmann S, Enzenbach C, Reineke A, Stausberg J, Damerow S, Huebner M, Schmidt B, Sauerbrei W, Richter A.
-Facilitating harmonized data quality assessments.
-A data quality framework for observational health research data collections with software implementations in R.
-BMC Med Res Meth 2021; 21(1): 1-15.
-[Link](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-021-01252-7)
-
-2.  *An overview of an R package to provide data quality assessments*
-
-Richter A , Schmidt CO , Krüger M , and Struckmann S.
-dataquieR: assessment of data quality in epidemiological research.
-Journal of Open Source Software 2021; 6(61), 3093, <https://doi.org/10.21105/joss.03093>
-
-3.  *Data management and data cleaning for a longitudinal survey design*
-
-Lusa L and Huebner M.
-Organizing and Analyzing Data from the SHARE Study with an Application to Age and Sex Differences in Depressive Symptoms.
-Int.
-J.
-Environ.
-Res.
-Public Health 2021, 18(18), 9684; <https://doi.org/10.3390/ijerph18189684>
-
-4.  *This is the accompanying R code for the IJEPRH paper (item 3)*
-
-Lusa L and Huebner M. Repository with R vignettes to support data organization and analysis for the SHARE study. 2021. <https://doi.org/10.17605/OSF.IO/KGTX6>
+<details>
+  <summary>Click for Answer</summary>
+    ```{r, echo=FALSE, eval=TRUE}
+    hist(1:10)
+    ```
+</details>

### 2025-04-16 papers.qmd
Test papers.qmd collapsible update

@@
   <summary>Click for Answer</summary>
-    ```{r, echo=FALSE, eval=TRUE}
-    hist(1:10)
-    ```
+Test Test
 </details>
+
+## State of the art in selection of variables and functional forms in multivariable analysis—outstanding issues
+
+Willi Sauerbrei,  Aris Perperoglou,  Matthias Schmid et al  for TG2 of the STRATOS initiative
+
+https://diagnprognres.biomedcentral.com/articles/10.1186/s41512-020-00074-3
+
+<details>
+  <summary>Abstract</summary>
+
+### Background
+
+How to select variables and identify functional forms for continuous variables is a key concern when creating a multivariable model.
+Ad hoc ‘traditional’ approaches to variable selection have been in use for at least 50 years.
+Similarly, methods for determining functional forms for continuous variables were first suggested many years ago.
+More recently, many alternative approaches to address these two challenges have been proposed, but knowledge of their properties and meaningful comparisons between them are scarce.
+To define a state of the art and to provide evidence-supported guidance to researchers who have only a basic level of statistical knowledge, many outstanding issues in multivariable modelling remain.
+Our main aims are to identify and illustrate such gaps in the literature and present them at a moderate technical level to the wide community of practitioners, researchers and students of statistics.
+
+### Methods
+
+We briefly discuss general issues in building descriptive regression models, strategies for variable selection, different ways of choosing functional forms for continuous variables and methods for combining the selection of variables and functions.
+We discuss two examples, taken from the medical literature, to illustrate problems in the practice of modelling.
+
+### Results
+
+Our overview revealed that there is not yet enough evidence on which to base recommendations for the selection of variables and functional forms in multivariable analysis.
+Such evidence may come from comparisons between alternative methods.
+In particular, we highlight seven important topics that require further investigation and make suggestions for the direction of further research.
+
+### Conclusions
+
+Selection of variables and of functional forms are important topics in multivariable analysis.
+To define a state of the art and to provide evidence-supported guidance to researchers who have only a basic level of statistical knowledge, further comparative research is required.
+</details>

### 2025-04-16 papers.qmd
Update papers.qmd with TG2 Content

@@
 
-https://diagnprognres.biomedcentral.com/articles/10.1186/s41512-020-00074-3
+[Link] (https://diagnprognres.biomedcentral.com/articles/10.1186/s41512-020-00074-3)
 
@@
 </details>
+
+## A review of spline function procedures in R
+
+Aris Perperoglou, Willi Sauerbrei, Michal Abrahamowicz, Matthias Schmid on behalf of TG2 of the STRATOS initiative
+
+[Link] (https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-019-0666-3)
+
+<details>
+  <summary>Abstract</summary>
+### Background
+
+With progress on both the theoretical and the computational fronts the use of spline modelling has become an established tool in statistical regression analysis.
+An important issue in spline modelling is the availability of user friendly, well documented software packages.
+Following the idea of the STRengthening Analytical Thinking for Observational Studies initiative to provide users with guidance documents on the application of statistical methods in observational research, the aim of this article is to provide an overview of the most widely used spline-based techniques and their implementation in R.
+
+### Methods
+
+In this work, we focus on the R Language for Statistical Computing which has become a hugely popular statistics software.
+We identified a set of packages that include functions for spline modelling within a regression framework.
+Using simulated and real data we provide an introduction to spline modelling and an overview of the most popular spline functions.
+
+### Results
+
+We present a series of simple scenarios of univariate data, where different basis functions are used to identify the correct functional form of an independent variable.
+Even in simple data, using routines from different packages would lead to different results.
+
+### Conclusions
+
+This work illustrate challenges that an analyst faces when working with data.
+Most differences can be attributed to the choice of hyper-parameters rather than the basis used.
+In fact an experienced user will know how to obtain a reasonable outcome, regardless of the type of spline used.
+However, many analysts do not have sufficient knowledge to use these powerful tools adequately and will need more guidance.
+</details>
+
+## Introducing the Topic Group on Selection of Variables and Functional Forms in Multivariable Analysis (TG2)
+
+Aris Perperoglou , Georg Heinze , Willi Sauerbrei on behalf of STRATOS TG2
+
+[Link] (https://stratos-initiative.org/sites/default/files/STRATOS-TG2-BB-variable-selection.pdf)
+
+<details>
+  <summary>Abstract</summary>
+The Biometric Bulletin has recently introduced its readership to the STRATOS initiative and described the activities of the Topic Groups on Missing Data (TG1), Measurement Error (TG4) and on Initial Data Analysis (TG3).
+This series now continues with an introduction to TG2, dealing with selection of variables and functional forms in multivariable analysis.
+</details>
+
+## Systematic review of education and practical guidance on regression modeling for medical researchers who lack a strong statistical background: Study protocol
+
+Paul Bach, Christine Wallisch, Nadja Klein, Lorena Hafermann, Willi Sauerbrei, Ewout W. Steyerberg, Georg Heinze, Geraldine Rauch , for topic group 2 of the STRATOS initiative
+
+[Link] (https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0241427)
+
+<details>
+  <summary>Abstract</summary>
+In the last decades, statistical methodology has developed rapidly, in particular in the field of regression modeling.
+Multivariable regression models are applied in almost all medical research projects.
+Therefore, the potential impact of statistical misconceptions within this field can be enormous Indeed, the current theoretical statistical knowledge is not always adequately transferred to the current practice in medical statistics.
+Some medical journals have identified this problem and published isolated statistical articles and even whole series thereof.
+In this systematic review, we aim to assess the current level of education on regression modeling that is provided to medical researchers via series of statistical articles published in medical journals.
+The present manuscript is a protocol for a systematic review that aims to assess which aspects of regression modeling are covered by statistical series published in medical journals that intend to train and guide applied medical researchers with limited statistical knowledge.
+Statistical paper series cannot easily be summarized and identified by common keywords in an electronic search engine like Scopus.
+We therefore identified series by a systematic request to statistical experts who are part or related to the STRATOS Initiative (STRengthening Analytical Thinking for Observational Studies).
+Within each identified article, two raters will independently check the content of the articles with respect to a predefined list of key aspects related to regression modeling.
+The content analysis of the topic-relevant articles will be performed using a predefined report form to assess the content as objectively as possible.
+Any disputes will be resolved by a third reviewer.
+Summary analyses will identify potential methodological gaps and misconceptions that may have an important impact on the quality of analyses in medical research.
+This review will thus provide a basis for future guidance papers and tutorials in the field of regression modeling which will enable medical researchers 1) to interpret publications in a correct way, 2) to perform basic statistical analyses in a correct way and 3) to identify situations when the help of a statistical expert is required.
+</details>
+
+# Sugggested Publications
+
+## Books
+
+* Harrell FE. Regression Modeling Strategies: With Applications to Linear Models, Logistic Regression, and Survival Analysis. Springer: New York, 2015.
+
+* Royston P, Sauerbrei W. Multivariable model-building: a pragmatic approach to regression analysis based on fractional polynomials for modelling continuous variables. John Wiley & Sons; 2008.
+
+* Wood S. Generalized Additive Models. Chapman & Hall/CRC: New York, 2006.
+
+* Miller A. Subset Selection in Regression. Taylor & Francis: Boca Raton, Florida, 2002.
+
+* Boer C de. A Practical Guide to Splines revised edn. Springer: New York, 2001.
+
+* Hastie T, Tibshirani R.. Generalized Additive Models. Chapman & Hall/CRC: New York, 1990.
+
+## Papers
+
+* Lu Z, Lou W.
+Bayesian approaches to variable selection: a comparative study from practical perspectives.
+Int J Biostat 2021; [https://doi.org/10.1515/ijb-2020-0130] (https://doi.org/10.1515/ijb-2020-0130)
+
+* Heinze G, Wallisch C, Dunkler D. Variable selection – a review and recommendations for the practicing statistician. Biometrical J. 2018; 60:431–49.
+
+* Binder H, Sauerbrei W, Royston P.
+Comparison between splines and fractional polynomials for multivariable model building with continuous covariates: a simulation study with continuous response.
+Statistics in Medicine 2013; 32(13): 2262– 2277.
+
+* Royston P, Sauerbrei W. Multivariable Model‐building. A Pragmatic Approach to Regression Analysis Based on Fractional Polynomials for Continuous Variables. Wiley: Chichester, 2008.
+
+* Sauerbrei W, Royston P, Binder H.
+Selection of important variables and determination of functional form for continuous predictors in multivariable model building.
+Statistics in Medicine 2007; 26: 5512– 5528.
+
+* Abrahamowicz M, MacKenzie TA. Joint estimation of time‐dependent and non‐linear effects of continuous covariates on survival. Statistics in Medicine 2007; 26(2): 392– 408.
+
+* Abrahamowicz M, Du Berger R, Grover SA. Flexible modeling of the effects of serum cholesterol on coronary heart disease mortality. American Journal of Epidemiology 1997; 145(8): 714– 729.
+
+* Greenland S. Avoiding power loss associated with categorization and ordinal scores in dose‐response and trend analysis. Epidemiology (Cambridge, Mass.) 1995; 6(4): 450– 454.
+
+* Royston P, Altman DG. Regression using fractional polynomials of continuous covariates: parsimonious parametric modelling. Applied Statistic 1994; 43(3): 429– 467.

### 2025-04-16 papers.qmd
Update papers.qmd with TG2 Content

@@
 ---
-
-<details>
-  <summary>Click for Answer</summary>
-Test Test
-</details>
 
@@
 
-[Link] (https://diagnprognres.biomedcentral.com/articles/10.1186/s41512-020-00074-3)
+[Link](https://diagnprognres.biomedcentral.com/articles/10.1186/s41512-020-00074-3)
 
@@
 
-[Link] (https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-019-0666-3)
+[Link](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-019-0666-3)
 
@@
 
-[Link] (https://stratos-initiative.org/sites/default/files/STRATOS-TG2-BB-variable-selection.pdf)
+[Link](https://stratos-initiative.org/sites/default/files/STRATOS-TG2-BB-variable-selection.pdf)
 
@@
 
-[Link] (https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0241427)
+[Link](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0241427)
 
@@
 Bayesian approaches to variable selection: a comparative study from practical perspectives.
-Int J Biostat 2021; [https://doi.org/10.1515/ijb-2020-0130] (https://doi.org/10.1515/ijb-2020-0130)
+Int J Biostat 2021; [https://doi.org/10.1515/ijb-2020-0130](https://doi.org/10.1515/ijb-2020-0130)
 

### 2025-04-16 resources.qmd
Update resources.qmd with TG2 content

@@
 ---
-title: "R implementations"
+title: "Software"
 ---
 
-### IDA report for regression models
+## Apps
 
-[Bacteremia data set](https://zenodo.org/records/7554815) by Georg Heinze
+**[Visualisation of non-linear modelling](https://biometrician.shinyapps.io/nonlinear_visualisation/): A shiny app developed by Daniela Dunkler and Georg Heinze**
 
-<https://stratosida.github.io/regression-regrets/>
+A shiny app to **visualize the possible different functional forms** which can be modelled with **fractional polynomials, natural (restricted cubic) splines, or linear B-splines.** For eleven different explanatory variables, you can visualise non-linear effects applying these popular methods of non-linear modeling.
+You can visualise first- or second-degree fractional polynomials, linear B-splines with up to 4 degrees of freedom and natural splines with up to 3 degrees of freedom.
 
-### IDA report for longitudinal studies
+## R packages on CRAN
 
-Grip strength data set from the [SHARE project](https://share-eric.eu)
+* Regression Modeling Strategies: [rms](https://cran.r-project.org/web/packages/rms/index.html) Regression modeling, testing, estimation, validation, graphics, prediction, and typesetting by storing enhanced model design attributes in the fit.
+'rms' is a collection of functions that assist with and streamline modeling.
 
-<https://stratosida.github.io/longitudinal/>
+* Augmented Backward Elimination: [abe](https://cran.r-project.org/web/packages/abe/index.html) Performs augmented backward elimination and checks the stability of the obtained model.
+Augmented backward elimination combines significance or information based criteria with the change in estimate to either select the optimal model for prediction purposes or to serve as a tool to obtain a practically sound, highly interpretable model.
+More details can be found in [Dunkler et al.
+(2014)](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0113677).
+
+* Multivariable Fractional Polynomials: [mfp](https://cran.r-project.org/web/packages/mfp/index.html)  Fractional polynomials are used to represent curvature in regression models.
+Software page: [https://mfp.imbi.uni-freiburg.de/software](https://mfp.imbi.uni-freiburg.de/software)

### 2025-04-17 glossary.qmd
Update glossary.qmd with TG2 contents

@@
 
-## Initial data analysis framework
+### Multivariable modelling
 
-### Initial Data Analysis
+Multivariable modeling in statistics is the process of using several independent variables to predict a single dependent variable.
+It accounts for the relationships between multiple variables and identifies which factors are most strongly associated with the outcome of interest.
+This method is commonly used in regression analysis and can help to improve the accuracy of predictions.
+
+[Perperoglou 2019](https://doi.org/10.1186/s12874-019-0666-3)
+
+### Functional form of continuous covariates
+
+The functional form of continuous covariates in statistics refers to the mathematical relationship between the independent variable and the dependent variable.
+This can be linear, nonlinear, or a combination of both.
+The choice of functional form can have a significant impact on the results of statistical analysis and should be carefully considered when selecting a model.
+### Spline function
+
+A spline function is a type of mathematical function used in statistics and numerical analysis to approximate complex curves or surfaces.
+It consists of multiple polynomial segments that join smoothly at specific points called knots.
+Spline functions are flexible and can be used to fit data or smooth curves.
+### R (software)
+
+R is a free and open-source software environment for statistical computing and graphics.
+It includes functions for data analysis, modeling, and visualization, and is extendable through packages.
+R is widely used by statisticians, data scientists, and researchers.
+
+### Regression
+
+Regression analysis is a statistical method used to investigate the relationship between a dependent variable and one or more independent variables.
+It can be used to model and predict outcomes based on input variables, and to identify which variables have the strongest impact on the outcome of interest.
+
+[Bach 2020](https://doi.org/10.1371/journal.pone.0241427)
+
+### Regression, Linear
+
+Linear regression is a statistical method used to model the relationship between a continuous dependent variable and one or more independent variables, which can be of any type.
+A linear function of the independent variables is used to model the expected value of the dependent variable.
+The task of estimating the model means to estimate the unknown coefficients of the independent variables.
+A constant (intercept) term is added to calibrate the predicted values of the dependent variable to have average zero errors (differences between predicted and observed values of the dependent variable, also called residuals).
+Mostly, but not necessarily, the coefficients are estimated by minimizing the sum of squared errors.
+Linear regression is commonly used for prediction and inference and is a fundamental tool in statistical analysis.
+
+### Regression, Logistic
+
+Logistic regression is a statistical method used to model the relationship between a binary dependent variable and one or more independent variables, which can be of any type.
+Unlike linear regression, the outcome variable in logistic regression is not continuous, but rather represents a binary outcome (e.g., success or failure).
+A linear function of the independent variables is used to model the log odds of one of the levels of the dependent variable.
+The task of estimating the model means to estimate the unknown coefficients of the independent variables.
+A constant (intercept) term is added to calibrate the predicted values of the dependent variable to have average zero errors.
+The coefficients are estimated by the maximum likelihood principle, which maximizes the probability of observing the given data given the model.
+Logistic regression is commonly used in classification problems, such as predicting whether a patient has a disease based on their symptoms.
+
+### Regression, Cox
+
+Cox regression, also known as proportional hazards regression, is a statistical method used to model the relationship between the time to an event and one or more independent variables, which can be of any type.
+The outcome variable is a time-to-event variable that can be censored in some observations (meaning the event of interest has not yet occurred).
+A linear function of the independent variables is used to model the log relative hazard of an event.
+The model assumes that the ratio of hazards between any two groups of individuals is constant over time (the proportional hazards assumption), but this assumption can be relaxed by specifying time-dependent coefficients.
+The coefficients are estimated by maximizing a partial likelihood, which accounts for the censoring of the data.
+Cox regression is commonly used in survival analysis and is a fundamental tool in medical research and epidemiology.
+
+### Regression, Poisson
+
+Poisson regression is a statistical method used to model the relationship between a count-type dependent variable and one or more independent variables, which can be of any type.
+A linear function of the independent variables is used to model the log of the expected count of the dependent variable.
+The model assumes that the expected count of the dependent variable is proportional to the value of the independent variables and that the counts are independent and identically distributed Poisson random variables.
+The coefficients of the model represent the change in the log expected count of the dependent variable associated with a one-unit change in the corresponding independent variable, holding all other variables constant.
+The coefficients are estimated by maximizing the likelihood of the model.
+Poisson regression is commonly used in fields such as epidemiology and finance to model count data.
+
+### Multivariable
+
+In statistical modeling, "multivariable" typically refers to models that include more than one independent variable (also known as predictor or explanatory variables) to predict a dependent variable.
+In contrast, "multivariate" refers to models that analyze multiple dependent variables at the same time, while "univariable" or "univariate" refers to models that only analyze one independent variable to predict a dependent variable.
+
+### Multivariate
+
+In statistical modeling, "multivariate" typically refers to models that analyze multiple dependent variables at the same time.
+This is in contrast to "multivariable," which refers to models that include more than one independent variable, and "univariate" or "univariable," which refers to models that only analyze one independent variable to predict a dependent variable.
+
+### Regression modeling
+
+See Regression
+
+### Descriptive modelling
+
+Descriptive modeling refers to a type of statistical modeling used to summarize a dataset and identify patterns, relationships, and trends without making any predictions or inferences.
+The purpose of descriptive models is to capture the main associations between a dependent variable and the independent variables in the data set, typically using measures such as means, medians, and standard deviations of target variables conditional on the values of other variables and associational quantities such as regression coefficients or nonlinear (smoothing) response functions.
+Descriptive models are useful for gaining insight into the underlying structure of the data, but are not intended for predictions of future outcomes or for causal conclusions.*
+
+[Sauerbrei 2020](https://doi.org/10.1186/s41512-020-00074-3)
+
+### Methods for variable selection
+
+Variable selection is the process of identifying the most relevant independent variables for a given dependent variable in a statistical model.
+One commonly used method is stepwise regression, which involves sequentially adding or removing variables based on their statistical significance.
+Another method is regularization, which shrinks the coefficients of less important variables towards zero and can be implemented through techniques such as Lasso or Ridge regression.
+Information criteria, such as AIC or BIC, can also be used to compare different models and select the best subset of variables.
+Finally, some machine learning algorithms, such as decision trees and random forests, automatically perform variable selection by choosing the most informative variables for predicting the outcome.
+
+### Spline procedures
+
+See Spline function
+
+### Fractional polynomials
+
+Fractional polynomials are a method for modeling nonlinear relationships between a continuous independent variable and a dependent variable in regression analysis.
+They involve transforming the independent variable by raising it to a power that is a fraction rather than a whole number, allowing for more flexible modeling of nonlinear relationships.
+Specifically, fractional polynomials use a series of power transformations (e.g., x^-2, x^-1, x^0.5, x^1, x^2) to estimate the functional form of the relationship between the variables.
+The best fitting model is selected by comparing the likelihoods of different fractional polynomial models, typically using the Akaike Information Criterion or the Bayesian Information Criterion.
+Fractional polynomials are useful for modeling complex nonlinear relationships between variables in a data-driven way.*
+
+### Categorisation
+
+Categorization, in the context of statistical modeling, is the process of grouping continuous independent variables into discrete categories based on a pre-specified set of cut-off points or thresholds.
+This is done in order to simplify the modeling process and to allow for the use of categorical variables in the model.
+Categorization can be used for a variety of reasons, including to reduce the effect of outliers or to capture nonlinear relationships between the independent and dependent variables.
+However, categorization can also lead to loss of information and can make the model less flexible.
+As such, the decision to categorize continuous variables should be based on careful consideration of the research question, the data, and the assumptions of the model.
+
+### Bias
+
+In statistical modeling, bias refers to the systematic error or deviation of a model's predicted values from the true values.
+Bias can arise for a variety of reasons, such as using an incorrect functional form for the model, selecting the wrong variables, or omitting important variables from the model.
+Bias can lead to underestimation or overestimation of the true relationship between the variables, and can result in inaccurate or misleading predictions.
+It is important to minimize bias in statistical modeling in order to obtain valid and reliable results.
+This can be achieved by using appropriate modeling techniques, careful variable selection, and by checking the assumptions of the model.
+Bias can also be reduced by using larger sample sizes, collecting better quality data, and by using appropriate statistical methods to correct for bias if it is detected.
+
+### Shrinkage
+
+In statistical modeling, shrinkage refers to the process of reducing the variability of model coefficients by constraining them towards a prior expectation or estimate.
+This is often done to avoid overfitting, which occurs when a model is too complex and fits the noise in the data rather than the underlying patterns or relationships.
+Shrinkage can be achieved through various techniques, such as ridge regression, lasso regression, and Bayesian modeling.
+These methods penalize large coefficients and encourage them to take on smaller values, thereby reducing the variance of the estimates.
+Shrinkage can improve the accuracy and generalizability of the model by reducing the effects of noise and improving the model's ability to make predictions on new data.
+
+### Empirical evidence
+
+In the context of the evaluation of statistical methods, empirical evidence refers to the results of experiments or analyses that provide insight into the performance of a given statistical method.
+Empirical evidence is based on the observed behavior of a method when applied to real-world data, and is used to determine whether the method is suitable for a particular task or problem.
+
+Empirical evidence can be obtained through various means, such as simulation studies, cross-validation, or experiments using real-world data.
+These studies are designed to assess the performance of a statistical method with respect to its accuracy, robustness, speed, and scalability.
+
+Empirical evidence is crucial in evaluating statistical methods because it provides a basis for determining their strengths and weaknesses, and for comparing their performance with that of other methods.
+The quality of the evidence depends on the design of the study, the choice of evaluation metrics, and the representativeness and relevance of the data.
+
+### Statistical regression modeling
+
+See Regression
+
+[Wallisch 2022](https://doi.org/10.1371/journal.pone.0262918)
+
+### Model performance
+
+Describes how accurate the predictions of a model are, but may also cover aspects of model interpretability.
+
+[Ullmann 2024](https://doi.org/10.1371/journal.pone.0308543)
+
+### Simulation study
+
+A simulation study is a research approach in which artificial datasets are generated under known conditions (using predefined statistical models and parameters) to systematically evaluate the performance of statistical methods.
+This allows researchers to assess how well different methods perform across a range of scenarios—such as varying sample sizes, noise levels, or model assumptions—when the "true" data-generating process is known.
+It's especially useful for comparing methods in terms of accuracy, bias, variance, and robustness.
+
+### Neutral comparison study
+
+A neutral comparison study is a type of research specifically designed to compare statistical methods in an unbiased and objective manner.
+In such studies, the primary goal is not to promote a new method but to fairly evaluate the relative performance of existing methods using consistent criteria and transparent protocols.
+Neutrality is maintained by ensuring that the study is not conducted by the developers of the methods being evaluated (or at least includes independent collaborators), and by using standardized, realistic datasets or simulations, along with clear, pre-specified performance metrics.
+These studies play a key role in evidence-based method selection and promote reproducibility and methodological rigor.
+
+### Initial data analysis
 
 The main aim of IDA is seen in providing reliable knowledge about the data to enable responsible statistical analyses and interpretation.
-IDA has the following phases: (1) metadata setup; (2) data cleaning; (3) data screening; (4) initial data reporting; 
-(5) refining and updating the research analysis plan; and (6) documenting and reporting IDA. IDA is aligned with the research aims 
-and the statistical analysis plan and does not include hypothesis generating activities.
+IDA has the following phases: (1) metadata setup; (2) data cleaning; (3) data screening; (4) initial data reporting; (5) refining and updating the research analysis plan; and (6) documenting and reporting IDA. IDA is aligned with the research aims and the statistical analysis plan and does not include hypothesis generating activities.
 
-**Source:** Schmidt CO, Vach W, le Cessie S, Huebner M. STRATOS: Introducing the Initial Data Analysis Topic Group (TG3). Biometric Bulletin 2018; 35 (2): 10-11
+**Source**: Schmidt CO, Vach W, le Cessie S, Huebner M. STRATOS: Introducing the Initial Data Analysis Topic Group (TG3). Biometric Bulletin 2018; 35 (2): 10-11
 
-[Link](https://www.stratos-initiative.org/sites/default/files/2021-11/STRATOS-TG3-BB-initial-data-analysis.pdf)
+[Heinze 2024](https://doi.org/10.1186/s12874-024-02294-3)
 
-### Exploratory Data Analysis
+### IDA framework
 
-Exploratory Data Analysis (EDA) inspects data to uncover patterns in the data or identifies possible errors and anomalies, for example through graphical approaches.
-EDA may include investigations for model building and hypothesis generation in a cyclical manner.
+A structured, principled approach to preparing and understanding data before formal statistical modeling begins.
+The IDA framework emphasizes systematic exploration and documentation of key aspects of the data, including data quality, distributions, missing values, outliers, and the plausibility of variable values.
+It also supports informed decisions about data cleaning, transformations, and the selection of analysis populations.
+In observational studies, the IDA framework helps ensure that subsequent analyses are based on a well-understood and appropriately prepared dataset, enhancing the credibility and transparency of research findings.
 
-**Source:** Lusa L, Proust-Lima C, Schmidt CO, Lee KJ, le Cessie S, Baillie M, Lawrence F, Huebner M; TG3 of the STRATOS Initiative.
-Initial data analysis for longitudinal studies to build a solid foundation for reproducible analysis.
-PLoS One.
-2024 May 29;19(5):e0295726.
-doi: 10.1371/journal.pone.0295726.
-
-[Link](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0295726)
-
-**References:**
-
-Behrens JT. Principles and procedures of exploratory data analysis. Psychological methods. 1997;2(2):131. <https://core.ac.uk/download/pdf/193648223.pdf>
-
-Cook D, Swayne DF. Interactive and Dynamic Graphics for Data Analysis: With R and Ggobi. New York, NY: Springer New York; 2007. <https://en.wikipedia.org/wiki/Exploratory_data_analysis>
-
-### Statistical Analysis Plan (SAP)
-
-A statistical analysis plan is a document that contains a more technical and detailed elaboration of the principal features of the analysis described in the protocol and includes detailed procedures for executing the statistical analysis of the primary and secondary variables and other data.
-
-**Source:** International Council for Harmonisation (ICH E9). ICH: E 9: Statistical principles for clinical trials - Step 5. Glossary p. 37
-
-[Link](https://www.ema.europa.eu/en/ich-e9-statistical-principles-clinical-trials-scientific-guideline)
-
-### Initial Data Analysis Plan (IDAP)
-
-An IDA plan is focused on data screening to prepare for statistical analyses; it does not specify procedures or software packages to conduct the analyses. It can be
-understood as a minimum basic set of analyses that can be extended depending on the context of the research study and data collection.
-IDA domains missing values, univariable and multivariable descriptions.
-
-**Source:** [Heinze et al https://doi.org/10.1186/s12874-024-02294-3;  Lusa et al https://doi.org/10.1371/journal.pone.0295726
-
-## Elements of initial data analysis
-
-### Data wrangling
-
-Data Wrangling is a broad term referring to the processes involved when preparing data for analysis.
-It can include acquiring data, enriching, changing the format and shape of the data, combining, subsetting and sampling data, and cleaning data. Some common steps involved with Data Wrangling are:
-
-* Discovering and gathering the data needed 
-* Merging data from different sources, if necessary  
-* Fixing flaws in the data entries  
-* Extracting the necessary data and put it in the proper structure  
-* Storing it in the proper format for further use
-
-**Source:** National Library of Medicine
-
-[Link](https://www.nnlm.gov/guides/data-glossary/data-wrangling)
-
-### Data cleaning
-
-Data cleaning is the process of identifying and correcting data that are inaccurate, missing, or incomplete.
-Data cleaning tasks can include removing duplicate records, investigating extreme values (e.g., outliers), converting dates from one format to another, removing unwanted text, splitting multiple data points in a cell into separate cells, or coding missing or NA values.
-
-**Source:** National Library of Medicine
-
-[Link](https://www.nnlm.gov/guides/data-glossary/data-cleaning)
+**Source**: Huebner M, le Cessie S, Schmidt CO, Vach W. A contemporary conceptual framework for initial data analysis. Observational Studies 2018; 4: 171-192.
 
@@
 
-**Source:** Huebner M, le Cessie S, Schmidt CO, Vach W. A contemporary conceptual framework for initial data analysis. Observational Studies 2018; 4: 171-192.
+**Source**: Huebner M, le Cessie S, Schmidt CO, Vach W. A contemporary conceptual framework for initial data analysis. Observational Studies 2018; 4: 171-192.
 
-[Link](https://doi.org/10.1353/obs.2018.0014)
+### Reporting
 
-### Structural variables
+The process of transparently and comprehensively documenting how an observational study was designed, conducted, analyzed, and interpreted.
+Good reporting ensures that all relevant details—such as data sources, inclusion criteria, variable definitions, statistical methods, handling of missing data, and sensitivity analyses—are clearly communicated.
+This enables readers to assess the validity, reproducibility, and generalizability of the findings.
+Adherence to established reporting guidelines (e.g., STROBE) is key to promoting clarity, transparency, and trust in observational research.
 
-Structural variables are variables that help to structure initial data analysis results for a clear organization and essential overview of data properties.
-Structuring can be based on levels of measurement (centers), on calendar time of recruitment, on demographic variables such as sex or age, or on variables of central importance to the research questions.
-Structural variables may or may not be included as predictors in the statistical models.
+### Functional form
 
-**Source:** [Heinze et al https://doi.org/10.1186/s12874-024-02294-3]
+See Functional forms of continuous covariates
 
-## Data and related terms
+### Variable transformation
 
-### Data
-
-Reinterpretable representation of information in a formalized manner suitable for communication, interpretation, or processing
-
-**Source:** ISO/IEC 2382:2015, 2121272, modified
-
-[Link](https://www.iso.org/obp/ui/#iso:std:iso:8000:-2:ed-2:v1:en)
-
-### Data set
-
-logically meaningful grouping of data
-
- **Source:**
-
-[Link](https://www.iso.org/obp/ui/#iso:std:iso:8000:-2:ed-2:v1:en)
-
-### Metadata
-
-data defining and describing other data
-
-**Source:** ISO/IEC 11179‑1:2015, 3.2.16
-
-[Link](https://www.iso.org/obp/ui/#iso:std:iso:8000:-2:ed-2:v1:en)
-
-### Data dictionary
-
-A data dictionary is a document that outlines the structure, content, and meaning of a given variable.
-This includes what type of data is being collected (e.g.
-free text, numerical, categorical or group data), the full wording of a question, what values are allowable (e.g.
-numeric ranges, multiple choice codes), and what those values mean (e.g.
-0 = no high blood pressure diagnosis, 1 = borderline high blood pressure, 2 = high blood pressure).
-A data dictionary is a critical tool for data analysis and reproducibility.
-
-**Source:** National Library of Medicine
-
-[Link-NLM](https://www.nnlm.gov/guides/data-glossary/data-dictionary)
-
-[Link-USGS](https://www.usgs.gov/data-management/data-dictionaries)
-
-### Data quality
-
-Data quality is the degree to which a set of characteristics of data fulfils requirements (=need or expectation that is stated, generally implied or obligatory)
-
-**Source:** ISO 9000:2015, 3.6.2  and  ISO 9000:2015, 3.6.4
-
-[Link1](https://www.iso.org/obp/ui/#iso:std:iso:8000:-2:ed-2:v1:en)
-
-[Link2](https://www.iso.org/obp/ui/#iso:std:iso:8000:-2:ed-2:v1:en)
-
-### Data subject
-
-A Data subject is the individual the personal data relates to.
-
-**Source:** General Data Protection Regulation (GDPR)
-
-[Link](https://www.edpb.europa.eu/system/files/2023-06/Key%20GDPR%20Definitions%20Infographic.pdf)
-
-### Data provenance
-
-The term “data provenance”, sometimes called “data lineage,” refers to a documented trail that accounts for the origin of a piece of data and where it has moved from to where it is presently.
-The purpose of data provenance is to tell researchers the origin, changes to, and details supporting the confidence or validity of research data.
-The concept of provenance guarantees that data creators are transparent about their work and where it came from and provides a chain of information where data can be tracked as researchers use other researchers’ data and adapt it for their own purposes.
-
-**Source:** National Library of Medicine
-
-[Link](https://www.nnlm.gov/guides/data-glossary/data-provenance)
+The process of applying mathematical functions to variables before including them in a regression model, with the aim of improving model fit, meeting statistical assumptions (e.g., linearity, normality), or enhancing interpretability.
+Common transformations include logarithmic, square root, and standardization (e.g., z-scores).
+In observational studies, variable transformation can help address issues such as skewed distributions, outliers, or non-linear relationships between predictors and outcomes.
+Careful documentation and justification of transformations are essential for transparency and reproducibility.

### 2025-04-24 papers.qmd
Update papers.qmd content with additional publications

@@
 
+## Evaluating variable selection methods for multivariable regression models: A simulation study protocol
+
+Ullmann, T., Heinze, G., Hafermann, L., Schilhart-Wallisch, C., Dunkler, D., for TG2 of the STRATOS initiative, 2024.
+Evaluating variable selection methods for multivariable regression models: A simulation study protocol.
+PLoS ONE 19, e0308543.
+[https://doi.org/10.1371/journal.pone.0308543](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0308543)
+
+<details>
+  <summary>Abstract</summary>
+
+### Abstract
+
+Researchers often perform data-driven variable selection when modeling the associations between an outcome and multiple independent variables in regression analysis.
+Variable selection may improve the interpretability, parsimony and/or predictive accuracy of a model.
+Yet variable selection can also have negative consequences, such as false exclusion of important variables or inclusion of noise variables, biased estimation of regression coefficients, underestimated standard errors and invalid confidence intervals, as well as model instability.
+While the potential advantages and disadvantages of variable selection have been discussed in the literature for decades, few large-scale simulation studies have neutrally compared data-driven variable selection methods with respect to their consequences for the resulting models.
+We present the protocol for a simulation study that will evaluate different variable selection methods: forward selection, stepwise forward selection, backward elimination, augmented backward elimination, univariable selection, univariable selection followed by backward elimination, and penalized likelihood approaches (Lasso, relaxed Lasso, adaptive Lasso).
+These methods will be compared with respect to false inclusion and/or exclusion of variables, consequences on bias and variance of the estimated regression coefficients, the validity of the confidence intervals for the coefficients, the accuracy of the estimated variable importance ranking, and the predictive performance of the selected models.
+We consider both linear and logistic regression in a low-dimensional setting (20 independent variables with 10 true predictors and 10 noise variables).
+The simulation will be based on real-world data from the National Health and Nutrition Examination Survey (NHANES).
+Publishing this study protocol ahead of performing the simulation increases transparency and allows integrating the perspective of other experts into the study design.
+
+</details>
+
+## Regression without regrets –initial data analysis is a prerequisite for multivariable regression
+
+Heinze, G., Baillie, M., Lusa, L., Sauerbrei, W., Schmidt, C.O., Harrell, F.E., Huebner, M., on behalf of TG2 and TG3 of the STRATOS initiative, 2024.
+Regression without regrets –initial data analysis is a prerequisite for multivariable regression.
+BMC Med Res Methodol 24, 178.
+[https://doi.org/10.1186/s12874-024-02294-3](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-024-02294-3)
+
+<details>
+  <summary>Abstract</summary>
+
+### Abstract
+
+Statistical regression models are used for predicting outcomes based on the values of some predictor variables or for describing the association of an outcome with predictors.
+With a data set at hand, a regression model can be easily fit with standard software packages.
+This bears the risk that data analysts may rush to perform sophisticated analyses without sufficient knowledge of basic properties, associations in and errors of their data, leading to wrong interpretation and presentation of the modeling results that lacks clarity.
+Ignorance about special features of the data such as redundancies or particular distributions may even invalidate the chosen analysis strategy.
+Initial data analysis (IDA) is prerequisite to regression analyses as it provides knowledge about the data needed to confirm the appropriateness of or to refine a chosen model building strategy, to interpret the modeling results correctly, and to guide the presentation of modeling results.
+In order to facilitate reproducibility, IDA needs to be preplanned, an IDA plan should be included in the general statistical analysis plan of a research project, and results should be well documented.
+Biased statistical inference of the final regression model can be minimized if IDA abstains from evaluating associations of outcome and predictors, a key principle of IDA.
+We give advice on which aspects to consider in an IDA plan for data screening in the context of regression modeling to supplement the statistical analysis plan.
+We illustrate this IDA plan for data screening in an example of a typical diagnostic modeling project and give recommendations for data visualizations.
+
+</details>
+
+## Review of guidance papers on regression modeling in statistical series of medical journals
+
+Wallisch, C., Bach, P., Hafermann, L., Klein, N., Sauerbrei, W., Steyerberg, E.W., Heinze, G., Rauch, G., on behalf of topic group 2 of the STRATOS initiative, 2022.
+Review of guidance papers on regression modeling in statistical series of medical journals.
+PLoS ONE 17, e0262918.
+[https://doi.org/10.1371/journal.pone.0262918](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0262918)
+
+### Abstract
+
+<details>
+  <summary>Abstract</summary>
+
+Although regression models play a central role in the analysis of medical research projects, there still exist many misconceptions on various aspects of modeling leading to faulty analyses.
+Indeed, the rapidly developing statistical methodology and its recent advances in regression modeling do not seem to be adequately reflected in many medical publications.
+This problem of knowledge transfer from statistical research to application was identified by some medical journals, which have published series of statistical tutorials and (shorter) papers mainly addressing medical researchers.
+The aim of this review was to assess the current level of knowledge with regard to regression modeling contained in such statistical papers.
+We searched for target series by a request to international statistical experts.
+We identified 23 series including 57 topic-relevant articles.
+Within each article, two independent raters analyzed the content by investigating 44 predefined aspects on regression modeling.
+We assessed to what extent the aspects were explained and if examples, software advices, and recommendations for or against specific methods were given.
+Most series (21/23) included at least one article on multivariable regression.
+Logistic regression was the most frequently described regression type (19/23), followed by linear regression (18/23), Cox regression and survival models (12/23) and Poisson regression (3/23).
+Most general aspects on regression modeling, e.g.
+model assumptions, reporting and interpretation of regression results, were covered.
+We did not find many misconceptions or misleading recommendations, but we identified relevant gaps, in particular with respect to addressing nonlinear effects of continuous predictors, model specification and variable selection.
+Specific recommendations on software were rarely given.
+Statistical guidance should be developed for nonlinear effects, model specification and variable selection to better support medical researchers who perform or interpret regression analyses.
+
+</details>
+
+## Systematic review of education and practical guidance on regression modeling for medical researchers who lack a strong statistical background: Study protocol
+
+Bach, P., Wallisch, C., Klein, N., Hafermann, L., Sauerbrei, W., Steyerberg, E.W., Heinze, G., Rauch, G., for topic group 2 of the STRATOS initiative, 2020.
+Systematic review of education and practical guidance on regression modeling for medical researchers who lack a strong statistical background: Study protocol.
+PLoS ONE 15, e0241427.
+[https://doi.org/10.1371/journal.pone.0241427](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0241427)
+
+<details>
+  <summary>Abstract</summary>
+In the last decades, statistical methodology has developed rapidly, in particular in the field of regression modeling.
+Multivariable regression models are applied in almost all medical research projects.
+Therefore, the potential impact of statistical misconceptions within this field can be enormous Indeed, the current theoretical statistical knowledge is not always adequately transferred to the current practice in medical statistics.
+Some medical journals have identified this problem and published isolated statistical articles and even whole series thereof.
+In this systematic review, we aim to assess the current level of education on regression modeling that is provided to medical researchers via series of statistical articles published in medical journals.
+The present manuscript is a protocol for a systematic review that aims to assess which aspects of regression modeling are covered by statistical series published in medical journals that intend to train and guide applied medical researchers with limited statistical knowledge.
+Statistical paper series cannot easily be summarized and identified by common keywords in an electronic search engine like Scopus.
+We therefore identified series by a systematic request to statistical experts who are part or related to the STRATOS Initiative (STRengthening Analytical Thinking for Observational Studies).
+Within each identified article, two raters will independently check the content of the articles with respect to a predefined list of key aspects related to regression modeling.
+The content analysis of the topic-relevant articles will be performed using a predefined report form to assess the content as objectively as possible.
+Any disputes will be resolved by a third reviewer.
+Summary analyses will identify potential methodological gaps and misconceptions that may have an important impact on the quality of analyses in medical research.
+This review will thus provide a basis for future guidance papers and tutorials in the field of regression modeling which will enable medical researchers 1) to interpret publications in a correct way, 2) to perform basic statistical analyses in a correct way and 3) to identify situations when the help of a statistical expert is required.
+</details>
+
 ## State of the art in selection of variables and functional forms in multivariable analysis—outstanding issues
 
-Willi Sauerbrei,  Aris Perperoglou,  Matthias Schmid et al  for TG2 of the STRATOS initiative
-
-[Link](https://diagnprognres.biomedcentral.com/articles/10.1186/s41512-020-00074-3)
+Sauerbrei, W., Perperoglou, A., Schmid, M., Abrahamowicz, M., Becher, H., Binder, H., Dunkler, D., Harrell, F.E., Royston, P., Heinze, G., for TG2 of the STRATOS initiative,  2020.
+State of the art in selection of variables and functional forms in multivariable analysis—outstanding issues.
+Diagn Progn Res 4, 3, s41512-020-00074–3.
+[https://doi.org/10.1186/s41512-020-00074-3](https://diagnprognres.biomedcentral.com/articles/10.1186/s41512-020-00074-3)
 
@@
 
-Aris Perperoglou, Willi Sauerbrei, Michal Abrahamowicz, Matthias Schmid on behalf of TG2 of the STRATOS initiative
-
-[Link](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-019-0666-3)
-
+Perperoglou, A., Sauerbrei, W., Abrahamowicz, M., Schmid, M., 2019.
+A review of spline function procedures in R.
+BMC Med Res Methodol 19, 46.
+[https://doi.org/10.1186/s12874-019-0666-3](https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-019-0666-3)
 <details>
@@
 
-[Link](https://stratos-initiative.org/sites/default/files/STRATOS-TG2-BB-variable-selection.pdf)
+[Link](https://www.stratos-initiative.org/sites/default/files/2021-11/STRATOS-TG2-BB-variable-selection.pdf)
 
@@
 This series now continues with an introduction to TG2, dealing with selection of variables and functional forms in multivariable analysis.
-</details>
-
-## Systematic review of education and practical guidance on regression modeling for medical researchers who lack a strong statistical background: Study protocol
-
-Paul Bach, Christine Wallisch, Nadja Klein, Lorena Hafermann, Willi Sauerbrei, Ewout W. Steyerberg, Georg Heinze, Geraldine Rauch , for topic group 2 of the STRATOS initiative
-
-[Link](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0241427)
-
-<details>
-  <summary>Abstract</summary>
-In the last decades, statistical methodology has developed rapidly, in particular in the field of regression modeling.
-Multivariable regression models are applied in almost all medical research projects.
-Therefore, the potential impact of statistical misconceptions within this field can be enormous Indeed, the current theoretical statistical knowledge is not always adequately transferred to the current practice in medical statistics.
-Some medical journals have identified this problem and published isolated statistical articles and even whole series thereof.
-In this systematic review, we aim to assess the current level of education on regression modeling that is provided to medical researchers via series of statistical articles published in medical journals.
-The present manuscript is a protocol for a systematic review that aims to assess which aspects of regression modeling are covered by statistical series published in medical journals that intend to train and guide applied medical researchers with limited statistical knowledge.
-Statistical paper series cannot easily be summarized and identified by common keywords in an electronic search engine like Scopus.
-We therefore identified series by a systematic request to statistical experts who are part or related to the STRATOS Initiative (STRengthening Analytical Thinking for Observational Studies).
-Within each identified article, two raters will independently check the content of the articles with respect to a predefined list of key aspects related to regression modeling.
-The content analysis of the topic-relevant articles will be performed using a predefined report form to assess the content as objectively as possible.
-Any disputes will be resolved by a third reviewer.
-Summary analyses will identify potential methodological gaps and misconceptions that may have an important impact on the quality of analyses in medical research.
-This review will thus provide a basis for future guidance papers and tutorials in the field of regression modeling which will enable medical researchers 1) to interpret publications in a correct way, 2) to perform basic statistical analyses in a correct way and 3) to identify situations when the help of a statistical expert is required.
 </details>

### 2025-05-02 about.qmd
Full update

@@
 ---
-title: "Members"
+title: "About this website"
 ---
 
-[Marianne Huebner](https://cstat.msu.edu/people/marianne-huebner-phd-sheher) (chair), Professor of Statistics and Probability at Michigan State University, USA. <https://orcid.org/0000-0002-9694-9231>
+**Andreas Klinger**: set-up and maintenance
 
-Carsten Oliver Schmidt (chair), Professor of the Institute of Community Medicine, University Medicine of Greifswald
+**Georg Heinze**: supervision and revision
 
-Mark Baillie, Advanced Methodology and Data Science, Clinical Development and Analytics. Novartis Pharma AG, Basel, Switzerland
+Last changes: 2 May 2025.
 
-Saskia le Cessie, Professor at Leiden University Medical Center, Netherlands
+### Acknowledgements
 
-[Lara Lusa](https://www.famnit.upr.si/sl/zaposleni-in-sodelavci/lara.lusa/), Professor at the Faculty of Mathematics, Natural Sciences and Information Technologies of the University of Primorska in Koper, Slovenia
+Many thanks to Marianne Huebner, who designed the website of the STRATOS Topic Group 3, which the website of Topic Group 2 is based on.
 
-[Michael Schell](https://www.moffitt.org/research-science/researchers/michael-schell/), Moffit Cancer Center, Florida, USA
+### Address for correspondence
 
-## **Bios**
-
-**Mark Baillie** is a statistician supporting the clinical development and analytics department at Novartis.
-He is a member of the methodology and consulting group, providing strategic and methodological input to various clinical and non-clinical priority projects across the company.
-He has a focus on effective visual communication working on a number of internal and external initiatives to improve the reporting of clinical trials and exploratory projects.
-
-**Marianne Huebner** is Professor of the Department of Statistics and Probability and Director of the Center for Statistical Training and Consulting at Michigan State University.
-After receiving a PhD in Applied Mathematics she held positions at UC Berkeley, the Mathematical Sciences Research Institute in Berkeley, the Mayo Clinic, and the University Medical Center in Hamburg, Germany.
-She develops and applies statistical models motivated by scientific questions.
-Her research interests include modeling health outcomes, initial data analyses, and sports and exercise.
-
-**Saskia le Cessie** is a statistician with a joint appointment at the Department of Clinical Epidemiology and the Section Medical Statistics, Department of Biomedical Datasciences.
-She obtained a master in Mathematics (with minor in Computer Sciences) at the University of Utrecht and obtained her PhD in Medical Statistics at the University of Leiden.
-She is a broadly oriented statistician and involved in several large epidemiological studies of the Department of Clinical Epidemiology.
-Her research interests are in epidemiological and statistical methods for observational studies, in particular instrumental variable analysis, mediation analysis, competing risks, causal modelling, meta analysis and repeated measurements.
-
-**Lara Lusa** is Professor of Statistics at the Faculty of Mathematics, Natural Sciences and Information Technologies of the University of Primorska, Slovenia.
-After receiving a PhD in Applied Statistics from the University of Florence, she worked at the Italian National Cancer Institute of Milan, at the Institute of Biostatistics and Medical Informatics of the University of Ljubljana and was a visiting fellow at the Biometric Research Branch of the National Cancer Institute in Bethesda.
-Her current research interests are related to the development and validation of predictive models in bio‐medicine, with particular attention to applications using high‐dimensional data and rare events.
-She is also interested in statistical programming, design of simulation studies, and teaching statistics.
-Beside the methodological research in statistics, she is also heavily involved as a statistical consultant to researchers in medicine.
-
-**Michael Schell**
-
-**Carsten Oliver Schmidt** is Professor at the Institute of Community Medicine, University Medicine of Greifswald.
-He is deputy head of the Department of SHIP/ Clinical-Epidemiological Research and leads the functional division "Quality in the Health Sciences".
-He is PI of several methods and research quality related projects, and speaker of the methods section of the German Society for Epidemiology.
-He is responsible for quality management procedures in population based observational health studies, has expertise in data linkage of primary and secondary data including registries, claims data, hospitals, and in the development of research infrastructures.
-Research topics include research methods, the study of subclinical and clinical disorders, incidental findings.
+STRATOS TG2 c/o Institute of Clinical Biometrics, center for Medical Data Science, Medical University of Vienna, Vienna, Austria. E-mail: stratos.tg2@gmail.com.

### 2025-05-02 activities.qmd
Full update

@@
 
-NOTE: This is the text that was originally on the TG2 website and it is severely outdated!
+## Recent
 
-* Georg Heinze and Christine Wallisch will give a STRATOS-TG2 workshop on 'Variable selection – a review and recommendations for the practicing statistician' at the IBS-ROeS conference in Salzburg, 6 September 2021
+* Georg Heinze and Theresa Ullmann presented a 3-hour Statistics in Practice Showcase session at the International Biometric Conference in Atlanta, GA, USA, December 2024.
+Find out more at [https://www.ibc2024.org](https://www.ibc2024.org/events/showcase).
+[The slides are available at OSF here](https://osf.io/a5gtk/).
 
-* Aris Perperoglou will host a session on "STRATOS initiative - on various issues in multivariable model building" at the RSS International Conference in Manchester,  6-9 September, with talks from Bianca de Stavola, Tim Morris, Willi Sauerbrei and Michal Abrahamowicz.
+## Upcoming
 
-* Georg Heinze and Mark Baillie will host a session on 'STRATOS initiative - Tools for guidance for analysis of observational studies' at the IBS-ROeS conference in Salzburg, 7-10 September 2021, featuring Rolf Groenwold from TG2 with a talk on ' Using videos and other media for guidance'
+* Internal TG2 meeting in Vienna, 7-9 May 2025 at the [Center for Medical Data Science](https://data-science.meduniwien.ac.at), including a public talk by Matthias Schmid on [Achieving explainable machine learning by functional decomposition of black-box models into explainable predictor effects](https://viennabiometrics.github.io/abstracts/20250805_WBS_Schmid.pdf)
+
+* Minisymposium at the International Society of Clinical Biostatistics (ISCB) conference in Basel, CH, 28 Aug 2025.
+Topic of the session: Statistical Research needs to improve – on the important roles of simulation studies and guidance for analysis.
+Speakers (among others): Georg Heinze, Willi Sauerbrei, Michal Abrahamowicz [Find out more](https://iscb2025.info/mini-symposia.html#mini-symposium-3-statistical-research-needs-to-improve-on-the-important-roles-of-simulation-studies-and-guidance-for-analysis)
+
+* Invited session at the [ROeS 2025](https://roes2025.at/index.html) (Conference of the Austro-Swiss Region of the International Biometric Society) in Graz, A, 14-18 Sep 2025.
+Topic of the session: Contrasting Bayesian and frequentist model building for descriptive research questions - a paired-design experiment of STRATOS-TG2.
+Invited speaker: Mariana Nold, further speakers for TG2: Michael Kammer, Georg Heinze, Aliaksandr Hubin
+
+* Topic-contributed paper session at Joint Statistical Meeting (JSM), 6 Aug 2025.
+Speaker for TG2: Marc Henrion.
+Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.

### 2025-05-02 members.qmd
Full update

--- 2025-04-16 members.qmd
@@
 
-[Georg Heinze](https://www.meduniwien.ac.at/researcher/georg_heinze) (chair): Institute of Clinical Biometrics, Center for Medical Statistics, Informatics and Intelligent Systems, Medical University of Vienna,  Austria.
+**[Georg Heinze](https://www.meduniwien.ac.at/researcher/georg_heinze) (chair)**: Institute of Clinical Biometrics, Center for Medical Data Science, Medical University of Vienna,  Austria.
 
@@
 
-**Heiko Becher**: Institute for Medical Biometry and Epidemiology, University Medical Center Hamburg-Eppendorf, Hamburg, Germany
-
 **Harald Binder**:  Institute for Medical Biometry and Medical Informatics, Medical Center - University of Freiburg, Germany
 
-**Daniela Dunkler**: Institute of Clinical Biometrics, Center for Medical Statistics, Informatics and Intelligent Systems, Medical University of Vienna,  Austria
+**Daniela Dunkler**: Institute of Clinical Biometrics, Center for Medical Data Science, Medical University of Vienna,  Austria
 
-**Rolf Groenwold**:  Leiden University Medical Centre, Leiden, The Netherlands
+**Frank Harrell**: School of Medicine, Vanderbilt University Medical Center, Nashville, TN, USA
 
-**Frank Harrell**: School of Medicine, Vanderbilt University Medical Center, USA
+**Marc Henrion**: Malawi Liverpool Wellcome Research Programme, Blantyre, Malawi and Department of Clinical Sciences, Liverpool School of Tropical Medicine, Liverpool, UK
 
-**Nadja Klein**: School of Business and Economics , Humboldt-University of Berlin, Germany
+**Michael Kammer**: Division of Nephrology, Department of Medicine III and Institute of Clinical Biometrics, Center for Medical Data Science, Medical University of Vienna,  Austria
 
-**Geraldine Rauch**:  Institute of Biometry and Clinical  Epidemiology, Charité – Universitätsmedizin Berlin , Germany
+**Mariana Nold**: Department of Sociology, Friedrich-Schiller University of Jena, Jena, Germany
 
-**Patrick Royston**: Department of Statistical Science, UCL, UK
-
-**Matthias Schmid**:  Institute for Medicine Biometry, Informatics and Epidemiology, Bonn, Germany
+**Matthias Schmid**:  Institute for Medicine Biometry, Informatics and Epidemiology, University of Bonn, Bonn, Germany
 
@@
 
-**Michael Kammer**: Institute of Clinical Biometrics, Center for Medical Statistics, Informatics and Intelligent Systems, Medical University of Vienna,  Austria
+**Gregor Buch**: Preventive Cardiology and Preventive Medicine, Department of Cardiology, University Medical Center of the Johannes Gutenberg University Mainz, Mainz, Germany
 
@@
 
-**Kim Luijken**:  Leiden University Medical Centre, Leiden, The Netherlands
-
-**Christine Wallisch**: Institute of Clinical Biometrics, Center for Medical Statistics, Informatics and Intelligent Systems, Medical University of Vienna,  Austria
+**Theresa Ullmann**: Institute of Clinical Biometrics, Center for Medical Data Science, Medical University of Vienna,  Austria

### 2025-05-02 resources.qmd
Full update

@@
 
-## R packages on CRAN
+## R packages and other ressources provided by TG2 members
 
-* Regression Modeling Strategies: [rms](https://cran.r-project.org/web/packages/rms/index.html) Regression modeling, testing, estimation, validation, graphics, prediction, and typesetting by storing enhanced model design attributes in the fit.
+### Regression Modeling Strategies
+
+Regression modeling, testing, estimation, validation, graphics, prediction, and typesetting by storing enhanced model design attributes in the fit.
 'rms' is a collection of functions that assist with and streamline modeling.
 
-* Augmented Backward Elimination: [abe](https://cran.r-project.org/web/packages/abe/index.html) Performs augmented backward elimination and checks the stability of the obtained model.
+* R package: [rms](https://cran.r-project.org/web/packages/rms/index.html) 
+* Online book companion: [https://hbiostat.org/doc/rms/book/](https://hbiostat.org/doc/rms/book/)
+
+### Multivariable Fractional Polynomials (MFP)
+
+MFP is a pragmatic procedure to create a multivariable model with the twin aims of selecting important variables and determining a suitable functional form for continuous predictors.
+
+Ressources:
+
+* R-package [mfp](https://cran.r-project.org/web/packages/mfp/index.html)
+* R-package [mfp2](https://cran.r-project.org/package=mfp2)
+* MFP website with explanations, examples, software, links, etc.: [https://mfp.imbi.uni-freiburg.de/software](https://mfp.imbi.uni-freiburg.de/software)
+
+### Augmented Backward Elimination
+
 Augmented backward elimination combines significance or information based criteria with the change in estimate to either select the optimal model for prediction purposes or to serve as a tool to obtain a practically sound, highly interpretable model.
@@
 
-* Multivariable Fractional Polynomials: [mfp](https://cran.r-project.org/web/packages/mfp/index.html)  Fractional polynomials are used to represent curvature in regression models.
-Software page: [https://mfp.imbi.uni-freiburg.de/software](https://mfp.imbi.uni-freiburg.de/software)
+* R package: [abe](https://cran.r-project.org/web/packages/abe/index.html) Performs augmented backward elimination and checks 
+the stability of the obtained model.
+
+### Tetris-plot
+
+The Tetris-plot is a tool to visualize the instability of models obtained by variable selection.
+
+* Github website with R package (development version): https://gsk-biostatistics.github.io/tetrisplot

### 2025-05-02 glossary.qmd
further small edits

@@
 ---
+
+On this page, you find descriptions of all keywords used in our papers.
+
+## Keywords used in [Perperoglou 2019](https://doi.org/10.1186/s12874-019-0666-3)
 
@@
 
-[Perperoglou 2019](https://doi.org/10.1186/s12874-019-0666-3)
-
 ### Functional form of continuous covariates
@@
 The choice of functional form can have a significant impact on the results of statistical analysis and should be carefully considered when selecting a model.
+
 ### Spline function
@@
 Spline functions are flexible and can be used to fit data or smooth curves.
+
 ### R (software)
@@
 
+## Keywords used in [Bach 2020](https://doi.org/10.1371/journal.pone.0241427)
+
 ### Regression
@@
 It can be used to model and predict outcomes based on input variables, and to identify which variables have the strongest impact on the outcome of interest.
-
-[Bach 2020](https://doi.org/10.1371/journal.pone.0241427)
 
@@
 
+## Keywords used in [Sauerbrei 2020](https://doi.org/10.1186/s41512-020-00074-3)
+
 ### Descriptive modelling
@@
 Descriptive models are useful for gaining insight into the underlying structure of the data, but are not intended for predictions of future outcomes or for causal conclusions.*
-
-[Sauerbrei 2020](https://doi.org/10.1186/s41512-020-00074-3)
 
@@
 
+## Keywords used in [Wallisch 2022](https://doi.org/10.1371/journal.pone.0262918)
+
 ### Statistical regression modeling
@@
 
-[Wallisch 2022](https://doi.org/10.1371/journal.pone.0262918)
+## Keywords used in [Ullmann 2024](https://doi.org/10.1371/journal.pone.0308543)
 
@@
 Describes how accurate the predictions of a model are, but may also cover aspects of model interpretability.
-
-[Ullmann 2024](https://doi.org/10.1371/journal.pone.0308543)
 
@@
 
+## Keywords used in [Heinze 2024](https://doi.org/10.1186/s12874-024-02294-3)
+
 ### Initial data analysis
@@
 The main aim of IDA is seen in providing reliable knowledge about the data to enable responsible statistical analyses and interpretation.
-IDA has the following phases: (1) metadata setup; (2) data cleaning; (3) data screening; (4) initial data reporting; (5) refining and updating the research analysis plan; and (6) documenting and reporting IDA. IDA is aligned with the research aims and the statistical analysis plan and does not include hypothesis generating activities.
+IDA has the following phases: (1) metadata setup; (2) data cleaning; (3) data screening; (4) initial data reporting; (5) refining and updating the research analysis plan; and (6) documenting and reporting IDA.
+IDA is aligned with the research aims and the statistical analysis plan and does not include hypothesis generating activities.
 
 **Source**: Schmidt CO, Vach W, le Cessie S, Huebner M. STRATOS: Introducing the Initial Data Analysis Topic Group (TG3). Biometric Bulletin 2018; 35 (2): 10-11
-
-[Heinze 2024](https://doi.org/10.1186/s12874-024-02294-3)
 

### 2025-05-02 index.qmd
Update index.qmd with acknowledgements
small changes

@@
 ---
-title: "Selection of Variables and Functional Forms in Multivariable Analysis"
+title: "Introduction"
 ---
@@
 
-Derive guidance for variable and function selection  in multivariable analysis.
+The main aim of STRATOS topic group 2 is to develop guidance for variable and functional form selection in multivariable analyses.
 
@@
 
-The main focus of TG2 is to identify influential variables and gain insight into their individual and joint relationship with the outcome.
-Two of the (interrelated) main challenges are selection of variables for inclusion in a multivariable explanatory model, and choice of the functional forms for continuous variables
+Multivariable models play a dominanting role in all kinds of empirical research, and they are used in addressing descriptive, predictive and explanatory research questions.
+
+The main focus of multivariable model building is to identify influential variables and to gain insight into their individual and joint relationship with an outcome variable.
+Two of the (interrelated) main challenges are selection of variables for inclusion in a multivariable model, and choice of the functional forms for continuous variables.
+
+On this website we list our activities, TG2 publications, further reading and resources, and the members of our topic group.
 
@@
 ![2nd General Meeting of the STRATOS members at the Banff International Research Station (BIRS), Banff, Canada.](assets/STRATOS_Banff_second_meeting.png){fig-align="center"}
+
+### Acknowledgements
+
+Many thanks to Marianne Huebner, who designed the website of the STRATOS Topic Group 3, which the website of Topic Group 2 is based on.

### 2025-05-02 index.qmd
small edit

@@
 
-On this website we list our activities, TG2 publications, further reading and resources, and the members of our topic group.
+On this website we list topic group specific activities, TG2 publications, further reading and resources, and the members of our topic group.
 
@@
 
+Find out more about STRATOS at its website: [https://www.stratos-initiative.org](https://www.stratos-initiative.org).
+
 ![2nd General Meeting of the STRATOS members at the Banff International Research Station (BIRS), Banff, Canada.](assets/STRATOS_Banff_second_meeting.png){fig-align="center"}
 
-### Acknowledgements
+## Acknowledgements
 

### 2025-07-07 activities.qmd
after TG2 meeting
Revision after checks from TU, 250707

@@
 
+* Internal TG2 meeting in Vienna, 7-9 May 2025 at the [Center for Medical Data Science](https://data-science.meduniwien.ac.at), including a public talk by Matthias Schmid on [Achieving explainable machine learning by functional decomposition of black-box models into explainable predictor effects](https://viennabiometrics.github.io/abstracts/20250805_WBS_Schmid.pdf)
+
 ## Upcoming
 
-* Internal TG2 meeting in Vienna, 7-9 May 2025 at the [Center for Medical Data Science](https://data-science.meduniwien.ac.at), including a public talk by Matthias Schmid on [Achieving explainable machine learning by functional decomposition of black-box models into explainable predictor effects](https://viennabiometrics.github.io/abstracts/20250805_WBS_Schmid.pdf)
+* Topic-contributed paper session at Joint Statistical Meeting (JSM), 6 Aug 2025.
+Speaker for TG2: Marc Henrion.
+Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.[Find out more](https://ww2.amstat.org/meetings/jsm/2025/)
+
+* Invited session at 70th Annual Meeting of the German Society for Medical Informatics, Biometrics and Epidemiology (Deutsche Gesellschaft für Medizinische Informatik, Biometrie und Epidemiologie e.
+V., GMDS) in Jena, Germany, 7-11 Sep 2025.
+Speaker for TG2: Gregor Buch.
+Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.[Find out more (in German)](https://gmds2025.de/)
 
@@
 
-* Topic-contributed paper session at Joint Statistical Meeting (JSM), 6 Aug 2025.
-Speaker for TG2: Marc Henrion.
-Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.
+* Invited talk at the [ROeS 2025](https://roes2025.at/index.html) (Conference of the Austro-Swiss Region of the International Biometric Society) in Graz, A, 14-18 Sep 2025.
+Topic of the session: Towards best practices in medical statistics: The need to implement innovative scientific solutions.
+Speaker: Theresa Ullmann.
+Topic of the talk: Neutral Comparisons Of Statistical Methods: Opportunities, Challenges, And Common Issues

### 2025-07-07 glossary.qmd
Revision after checks from TU, 250707

@@
 
-### Multivariable modelling
+### Multivariable modeling
 
@@
 
-### Descriptive modelling
+### Descriptive modeling
 
-Descriptive modeling refers to a type of statistical modeling used to summarize a dataset and identify patterns, relationships, and trends without making any predictions or inferences.
+Descriptive modeling refers to a type of statistical modeling used to summarize a dataset and identify patterns, relationships, and trends.
 The purpose of descriptive models is to capture the main associations between a dependent variable and the independent variables in the data set, typically using measures such as means, medians, and standard deviations of target variables conditional on the values of other variables and associational quantities such as regression coefficients or nonlinear (smoothing) response functions.
-Descriptive models are useful for gaining insight into the underlying structure of the data, but are not intended for predictions of future outcomes or for causal conclusions.*
+Descriptive models are useful for gaining insight into the underlying structure of the data, but are not mainly intended for predictions of future outcomes or for causal conclusions.
 
@@
 One commonly used method is stepwise regression, which involves sequentially adding or removing variables based on their statistical significance.
-Another method is regularization, which shrinks the coefficients of less important variables towards zero and can be implemented through techniques such as Lasso or Ridge regression.
+Another method is regularization, which shrinks the coefficients of less important variables towards zero and can be implemented through techniques such as Lasso or the Elastic Net.
 Information criteria, such as AIC or BIC, can also be used to compare different models and select the best subset of variables.

### 2025-07-07 index.qmd
Revision after checks from TU, 250707

@@
 
-Multivariable models play a dominanting role in all kinds of empirical research, and they are used in addressing descriptive, predictive and explanatory research questions.
+Multivariable models play a dominant role in all kinds of empirical research, and they are used in addressing descriptive, predictive and explanatory research questions.
 

### 2025-07-07 resources.qmd
updated link
further small edits
typo corrected
Revision after checks from TU, 250707

@@
 ---
-title: "Software"
+title: "Resources"
 ---
@@
 
-**[Visualisation of non-linear modelling](https://biometrician.shinyapps.io/nonlinear_visualisation/): A shiny app developed by Daniela Dunkler and Georg Heinze**
+**[Bend your (sp)line](https://clinicalbiometrics.shinyapps.io/Bendyourspline/): A shiny app developed by TG2**
 
@@
 
-## R packages and other ressources provided by TG2 members
+## R packages and other resources provided by TG2 members
 
@@
 
-Ressources:
+Resources:
 
@@
 
-* Github website with R package (development version): https://gsk-biostatistics.github.io/tetrisplot
+* Github website with R package (development version): [https://gsk-biostatistics.github.io/tetrisplot](https://gsk-biostatistics.github.io/tetrisplot)

### 2025-08-11 activities.qmd
Revision 250707, small amendment
Minor updates by Aug 11, 2025 (GH)

@@
 
+* Topic-contributed paper session at Joint Statistical Meeting (JSM), 6 Aug 2025.
+Speaker for TG2: Marc Henrion.
+Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.[Find out more](https://ww2.amstat.org/meetings/jsm/2025/)
+
+* Internal TG2 meeting in Vienna, 7-9 May 2025 at the [Center for Medical Data Science](https://data-science.meduniwien.ac.at), including a public talk by Matthias Schmid on [Achieving explainable machine learning by functional decomposition of black-box models into explainable predictor effects](https://viennabiometrics.github.io/abstracts/20250805_WBS_Schmid.pdf)
+
 * Georg Heinze and Theresa Ullmann presented a 3-hour Statistics in Practice Showcase session at the International Biometric Conference in Atlanta, GA, USA, December 2024.
@@
 
-* Internal TG2 meeting in Vienna, 7-9 May 2025 at the [Center for Medical Data Science](https://data-science.meduniwien.ac.at), including a public talk by Matthias Schmid on [Achieving explainable machine learning by functional decomposition of black-box models into explainable predictor effects](https://viennabiometrics.github.io/abstracts/20250805_WBS_Schmid.pdf)
-
 ## Upcoming
 
-* Topic-contributed paper session at Joint Statistical Meeting (JSM), 6 Aug 2025.
-Speaker for TG2: Marc Henrion.
-Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.[Find out more](https://ww2.amstat.org/meetings/jsm/2025/)
+* Minisymposium at the International Society of Clinical Biostatistics (ISCB) conference in Basel, CH, 28 Aug 2025.
+Topic of the session: Statistical Research needs to improve – on the important roles of simulation studies and guidance for analysis.
+Speakers (among others): Georg Heinze, Willi Sauerbrei, Michal Abrahamowicz.
+[Find out more](https://iscb2025.info/mini-symposia.html#mini-symposium-3-statistical-research-needs-to-improve-on-the-important-roles-of-simulation-studies-and-guidance-for-analysis)
 
@@
 Speaker for TG2: Gregor Buch.
-Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.[Find out more (in German)](https://gmds2025.de/)
-
-* Minisymposium at the International Society of Clinical Biostatistics (ISCB) conference in Basel, CH, 28 Aug 2025.
-Topic of the session: Statistical Research needs to improve – on the important roles of simulation studies and guidance for analysis.
-Speakers (among others): Georg Heinze, Willi Sauerbrei, Michal Abrahamowicz [Find out more](https://iscb2025.info/mini-symposia.html#mini-symposium-3-statistical-research-needs-to-improve-on-the-important-roles-of-simulation-studies-and-guidance-for-analysis)
+Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.
+[Find out more (in German)](https://gmds2025.de/)
 
@@
 Speaker: Theresa Ullmann.
-Topic of the talk: Neutral Comparisons Of Statistical Methods: Opportunities, Challenges, And Common Issues
+Topic of the talk: Neutral Comparisons Of Statistical Methods: Opportunities, Challenges, And Common Issues (Includes as an example TG2-P11, Simulation study on variable selection)

### 2025-08-11 index.qmd
Minor updates by Aug 11, 2025 (GH)

@@
 
-![2nd General Meeting of the STRATOS members at the Banff International Research Station (BIRS), Banff, Canada.](assets/STRATOS_Banff_second_meeting.png){fig-align="center"}
+![2nd General Meeting of the STRATOS members at the Banff International Research Station (BIRS), Banff, Canada, June 2019.](assets/STRATOS_Banff_second_meeting.png){fig-align="center"}
 

### 2025-08-11 members.qmd
Corrected affiliation
added CSL
Minor updates by Aug 11, 2025 (GH)

@@
 
-**Aris Perperoglou (chair)**:  Data Science & AI,   AstraZeneca R&D Biopharmaceuticals , Cambridge UK
+**[Aris Perperoglou](https://uk.linkedin.com/in/arisperperoglou) (chair)**:  Predictive Modelling, GSK, Stevenage, UK
 
-**Willi Sauerbrei (chair)**: Institute for Medical Biometry and Medical Informatics, Medical Center - University of Freiburg, Germany
+**[Willi Sauerbrei](https://www.uniklinik-freiburg.de/english/imbi/mitarbeiter.html?imbiuser=wfs) (chair)**: Institute for Medical Biometry and Medical Informatics, Medical Center - University of Freiburg, Germany
 
@@
 
-**Daniela Dunkler**: Institute of Clinical Biometrics, Center for Medical Data Science, Medical University of Vienna,  Austria
+**[Daniela Dunkler](https://www.meduniwien.ac.at/researcher/daniela_dunkler)**: Institute of Clinical Biometrics, Center for Medical Data Science, Medical University of Vienna,  Austria
 
@@
 
-**Mariana Nold**: Department of Sociology, Friedrich-Schiller University of Jena, Jena, Germany
+**[Mariana Nold](https://www.fsv.uni-jena.de/19575/dr-mariana-nold)**: Department of Sociology, Friedrich-Schiller University of Jena, Jena, Germany
 
 **Matthias Schmid**:  Institute for Medicine Biometry, Informatics and Epidemiology, University of Bonn, Bonn, Germany
+
+## Clinical affiliate
+
+**Carl-Stephan Leonhardt**: Department of Surgery, Medical University of Vienna, Vienna, Austria
 
@@
 
-**Theresa Ullmann**: Institute of Clinical Biometrics, Center for Medical Data Science, Medical University of Vienna,  Austria
+**[Theresa Ullmann](https://www.meduniwien.ac.at/researcher/theresa_ullmann)**: Institute of Clinical Biometrics, Center for Medical Data Science, Medical University of Vienna,  Austria

### 2025-08-11 papers.qmd
Test update on papers.qmd
Full update
Minor updates by Aug 11, 2025 (GH)

@@
 
+## Recent activities of the Topic Group on Selection of Variables and Functional Forms in Multivariable Analysis (TG2)
+
+Georg Heinze, Aris Perperoglou, Willi Sauerbrei on behalf of STRATOS TG2
+
+[Biometric Bulletin 2021; 38(2):7-8](https://www.stratos-initiative.org/sites/default/files/2021-09/IBS_BioMetric-Bulletin-Vol38-Issue2-1_4-7-8.pdf)
+
+<details>
+<summary>Abstract</summary>
+This Biometric Bulletin article provides an update of the activities of TG2 by July 2021.
+</details>
+
 ## Introducing the Topic Group on Selection of Variables and Functional Forms in Multivariable Analysis (TG2)
 
-Aris Perperoglou , Georg Heinze , Willi Sauerbrei on behalf of STRATOS TG2
+Aris Perperoglou, Georg Heinze, Willi Sauerbrei on behalf of STRATOS TG2
 
-[Link](https://www.stratos-initiative.org/sites/default/files/2021-11/STRATOS-TG2-BB-variable-selection.pdf)
+[Biometric Bulletin 2018; 35(3):18-19](https://www.stratos-initiative.org/sites/default/files/2021-11/STRATOS-TG2-BB-variable-selection.pdf)
 
@@
 
-# Sugggested Publications
+# Suggested Publications
 
@@
 
+* Kipruto, E., & Sauerbrei, W.
+(2025).
+Evaluating Prediction Performance: A Simulation Study Comparing Penalized and Classical Variable Selection Methods in Low-Dimensional Data.
+Applied Sciences, 15(13), 7443.
+[https://doi.org/10.3390/app15137443](https://doi.org/10.3390/app15137443)
+
+* Sauerbrei, W., Royston, P., & Kipruto, E.
+(2025).
+Multivariable Fractional Polynomial Models and Extensions.
+In International Encyclopedia of Statistical Science (pp.
+1609-1616).
+Berlin, Heidelberg: Springer.
+[https://doi.org/10.1007/978-3-662-69359-9_397](https://doi.org/10.1007/978-3-662-69359-9_397)
+
+* Kipruto, E., & Sauerbrei, W.
+(2024).
+Post‐Estimation Shrinkage in Full and Selected Linear Regression Models in Low‐Dimensional Data Revisited.
+Biometrical Journal, 66(7), e202300368.
+[https://doi.org/10.1002/bimj.202300368](https://doi.org/10.1002/bimj.202300368)
+
 * Lu Z, Lou W.
@@
 * Heinze G, Wallisch C, Dunkler D. Variable selection – a review and recommendations for the practicing statistician. Biometrical J. 2018; 60:431–49. 
-
 * Binder H, Sauerbrei W, Royston P.

### 2025-08-12 members.qmd
minor edit, 250812GH

@@
 
-**[Willi Sauerbrei](https://www.uniklinik-freiburg.de/english/imbi/mitarbeiter.html?imbiuser=wfs) (chair)**: Institute for Medical Biometry and Medical Informatics, Medical Center - University of Freiburg, Germany
+**[Willi Sauerbrei](https://www.uniklinik-freiburg.de/imbi-en/employees.html?imbiuser=wfs) (chair)**: Institute for Medical Biometry and Medical Informatics, Medical Center - University of Freiburg, Germany
 

### 2025-08-18 papers.qmd
minor edit, 250812GH
Added short sentence describing why the books are relevant

@@
 
-* Harrell FE. Regression Modeling Strategies: With Applications to Linear Models, Logistic Regression, and Survival Analysis. Springer: New York, 2015.
+* Harrell FE.
+Regression Modeling Strategies: With Applications to Linear Models, Logistic Regression, and Survival Analysis.
+Springer: New York, 2015.
+<br> *Integrates statistical rigor with practical guidance on variable selection, model specification, and validation across a range of regression contexts*.
 
-* Royston P, Sauerbrei W. Multivariable model-building: a pragmatic approach to regression analysis based on fractional polynomials for modelling continuous variables. John Wiley & Sons; 2008.
+* Royston P, Sauerbrei W.
+Multivariable model-building: a pragmatic approach to regression analysis based on fractional polynomials for modelling continuous variables.
+John Wiley & Sons; 2008.
+<br> *Presents a structured approach to modeling continuous variables using fractional polynomials, with an emphasis on reproducibility and clinical relevance*.
 
-* Wood S. Generalized Additive Models. Chapman & Hall/CRC: New York, 2006.
+* Wood S.
+Generalized Additive Models.
+Chapman & Hall/CRC: New York, 2006.
+<br> *Offers an accessible yet thorough treatment of additive modeling techniques for capturing nonlinear relationships in multivariable settings*.
 
-* Miller A. Subset Selection in Regression. Taylor & Francis: Boca Raton, Florida, 2002.
+* Miller A.
+Subset Selection in Regression.
+Taylor & Francis: Boca Raton, Florida, 2002.
+<br> *Provides a critical examination of subset selection methods, discussing their statistical properties, limitations, and applications*.
 
-* Boer C de. A Practical Guide to Splines revised edn. Springer: New York, 2001.
+* Boer C de.
+A Practical Guide to Splines revised edn.
+Springer: New York, 2001.
+<br> *Covers spline-based methods in detail, offering practical advice for incorporating flexible functional forms into regression models*.
 
-* Hastie T, Tibshirani R.. Generalized Additive Models. Chapman & Hall/CRC: New York, 1990.
+* Hastie T, Tibshirani R..
+Generalized Additive Models.
+Chapman & Hall/CRC: New York, 1990.
+<br> *Introduces foundational concepts in additive modeling, balancing statistical theory with illustrative applied examples*.
 
@@
 
-* Royston P, Sauerbrei W. Multivariable Model‐building. A Pragmatic Approach to Regression Analysis Based on Fractional Polynomials for Continuous Variables. Wiley: Chichester, 2008.
-
 * Sauerbrei W, Royston P, Binder H.

### 2025-08-18 papers.qmd
Added short sentence explaining why it is useful for each paper.

@@
 [https://doi.org/10.3390/app15137443](https://doi.org/10.3390/app15137443)
+<br> *Highlights how the relative performance of penalized and classical selection methods depends on data conditions, guiding informed method choice*.
 
@@
 [https://doi.org/10.1007/978-3-662-69359-9_397](https://doi.org/10.1007/978-3-662-69359-9_397)
+<br> *Summarizes fractional polynomial modeling as a structured and practical approach to functional form selection.*
 
@@
 [https://doi.org/10.1002/bimj.202300368](https://doi.org/10.1002/bimj.202300368)
+<br> *Demonstrates how post-estimation shrinkage can reduce overfitting and bias, offering an alternative to penalized regression after variable selection*.
 
@@
 Int J Biostat 2021; [https://doi.org/10.1515/ijb-2020-0130](https://doi.org/10.1515/ijb-2020-0130)
+<br> *Provides practical comparisons of Bayesian variable selection approaches, supporting their application in real-world data analysis*.
 
 * Heinze G, Wallisch C, Dunkler D. Variable selection – a review and recommendations for the practicing statistician. Biometrical J. 2018; 60:431–49. 
+<br> *Offers pragmatic recommendations on variable selection methods, emphasizing model stability, validity, and transparency*.
+
 * Binder H, Sauerbrei W, Royston P.
@@
 Statistics in Medicine 2013; 32(13): 2262– 2277.
+<br> *Compares splines and fractional polynomials, clarifying their relative strengths for modeling continuous covariates*.
 
@@
 Statistics in Medicine 2007; 26: 5512– 5528.
+<br> *Integrates strategies for variable selection and functional form choice, presenting fractional polynomials as a coherent solution*.
 
 * Abrahamowicz M, MacKenzie TA. Joint estimation of time‐dependent and non‐linear effects of continuous covariates on survival. Statistics in Medicine 2007; 26(2): 392– 408.
+<br> *Illustrates how joint modeling of time-varying and nonlinear effects improves accuracy and avoids bias in survival analysis*.
 
 * Abrahamowicz M, Du Berger R, Grover SA. Flexible modeling of the effects of serum cholesterol on coronary heart disease mortality. American Journal of Epidemiology 1997; 145(8): 714– 729.
+<br> *Shows how flexible modeling of continuous exposures can reveal nonlinear risk relationships that parametric models may obscure*.
 
 * Greenland S. Avoiding power loss associated with categorization and ordinal scores in dose‐response and trend analysis. Epidemiology (Cambridge, Mass.) 1995; 6(4): 450– 454.
+<br> *Demonstrates the loss of power and information from categorizing continuous exposures, reinforcing the need for flexible functional form modeling*.
 
 * Royston P, Altman DG. Regression using fractional polynomials of continuous covariates: parsimonious parametric modelling. Applied Statistic 1994; 43(3): 429– 467. 
+<br> *Introduces fractional polynomials as a parsimonious and interpretable alternative to categorization or high-order polynomials*.

### 2025-08-18 papers.qmd
Added DOI to all papers

@@
 
-* Heinze G, Wallisch C, Dunkler D. Variable selection – a review and recommendations for the practicing statistician. Biometrical J. 2018; 60:431–49. 
+* Heinze G, Wallisch C, Dunkler D.
+Variable selection – a review and recommendations for the practicing statistician.
+Biometrical J.
+2018; 60:431–49.
+[https://doi.org/10.1002/bimj.201700067](https://doi.org/10.1002/bimj.201700067)
 <br> *Offers pragmatic recommendations on variable selection methods, emphasizing model stability, validity, and transparency*.
@@
 Statistics in Medicine 2013; 32(13): 2262– 2277.
+[https://doi.org/10.1002/sim.5639](https://doi.org/10.1002/sim.5639)
 <br> *Compares splines and fractional polynomials, clarifying their relative strengths for modeling continuous covariates*.
@@
 Statistics in Medicine 2007; 26: 5512– 5528.
+[https://doi.org/10.1002/sim.3148](https://doi.org/10.1002/sim.3148)
 <br> *Integrates strategies for variable selection and functional form choice, presenting fractional polynomials as a coherent solution*.
 
-* Abrahamowicz M, MacKenzie TA. Joint estimation of time‐dependent and non‐linear effects of continuous covariates on survival. Statistics in Medicine 2007; 26(2): 392– 408.
+* Abrahamowicz M, MacKenzie TA.
+Joint estimation of time‐dependent and non‐linear effects of continuous covariates on survival.
+Statistics in Medicine 2007; 26(2): 392– 408.
+[https://doi.org/10.1002/sim.2519](https://doi.org/10.1002/sim.2519)
 <br> *Illustrates how joint modeling of time-varying and nonlinear effects improves accuracy and avoids bias in survival analysis*.
 
-* Abrahamowicz M, Du Berger R, Grover SA. Flexible modeling of the effects of serum cholesterol on coronary heart disease mortality. American Journal of Epidemiology 1997; 145(8): 714– 729.
+* Abrahamowicz M, Du Berger R, Grover SA.
+Flexible modeling of the effects of serum cholesterol on coronary heart disease mortality.
+American Journal of Epidemiology 1997; 145(8): 714– 729.
+[https://doi.org/10.1093/aje/145.8.714](https://doi.org/10.1093/aje/145.8.714)
 <br> *Shows how flexible modeling of continuous exposures can reveal nonlinear risk relationships that parametric models may obscure*.
 
-* Greenland S. Avoiding power loss associated with categorization and ordinal scores in dose‐response and trend analysis. Epidemiology (Cambridge, Mass.) 1995; 6(4): 450– 454.
+* Greenland S.
+Avoiding power loss associated with categorization and ordinal scores in dose‐response and trend analysis.
+Epidemiology (Cambridge, Mass.) 1995; 6(4): 450– 454.
+[https://doi.org/10.1097/00001648-199507000-00025](https://doi.org/10.1097/00001648-199507000-00025)
 <br> *Demonstrates the loss of power and information from categorizing continuous exposures, reinforcing the need for flexible functional form modeling*.
 
-* Royston P, Altman DG. Regression using fractional polynomials of continuous covariates: parsimonious parametric modelling. Applied Statistic 1994; 43(3): 429– 467. 
+* Royston P, Altman DG.
+Regression using fractional polynomials of continuous covariates: parsimonious parametric modelling.
+Applied Statistic 1994; 43(3): 429– 467.
+[https://doi.org/10.2307/2986270](https://doi.org/10.2307/2986270)
 <br> *Introduces fractional polynomials as a parsimonious and interpretable alternative to categorization or high-order polynomials*.

### 2025-08-18 papers.qmd
Added DOI (or other link) for books

@@
 Springer: New York, 2015.
+[https://doi.org/10.1007/978-3-319-19425-7](https://doi.org/10.1007/978-3-319-19425-7)
 <br> *Integrates statistical rigor with practical guidance on variable selection, model specification, and validation across a range of regression contexts*.
@@
 John Wiley & Sons; 2008.
+[https://www.wiley.com/en-us/Multivariable+Model+-+Building%3A+A+Pragmatic+Approach+to+Regression+Anaylsis+based+on+Fractional+Polynomials+for+Modelling+Continuous+Variables-p-9780470028421](Link)
 <br> *Presents a structured approach to modeling continuous variables using fractional polynomials, with an emphasis on reproducibility and clinical relevance*.
 
-* Wood S.
-Generalized Additive Models.
-Chapman & Hall/CRC: New York, 2006.
+* Wood S. Generalized Additive Models. Chapman & Hall/CRC: New York, 2006. [https://doi.org/10.1201/9781315370279](https://doi.org/10.1201/9781315370279)
 <br> *Offers an accessible yet thorough treatment of additive modeling techniques for capturing nonlinear relationships in multivariable settings*.
 
-* Miller A.
-Subset Selection in Regression.
-Taylor & Francis: Boca Raton, Florida, 2002.
+* Miller A. Subset Selection in Regression. Taylor & Francis: Boca Raton, Florida, 2002. [https://doi.org/10.1201/9781420035933](https://doi.org/10.1201/9781420035933)
 <br> *Provides a critical examination of subset selection methods, discussing their statistical properties, limitations, and applications*.
 
-* Boer C de.
-A Practical Guide to Splines revised edn.
-Springer: New York, 2001.
+* Boer C de. A Practical Guide to Splines revised edn. Springer: New York, 2001. [https://link.springer.com/book/9780387953663](Link)
 <br> *Covers spline-based methods in detail, offering practical advice for incorporating flexible functional forms into regression models*.
 
-* Hastie T, Tibshirani R..
-Generalized Additive Models.
-Chapman & Hall/CRC: New York, 1990.
+* Hastie T, Tibshirani R.. Generalized Additive Models. Chapman & Hall/CRC: New York, 1990. [https://www.routledge.com/Generalized-Additive-Models/Hastie-Tibshirani/p/book/9780412343902](Link)
 <br> *Introduces foundational concepts in additive modeling, balancing statistical theory with illustrative applied examples*.

### 2025-08-18 papers.qmd
Small change in sentences for first two recommended books.

@@
 [https://doi.org/10.1007/978-3-319-19425-7](https://doi.org/10.1007/978-3-319-19425-7)
-<br> *Integrates statistical rigor with practical guidance on variable selection, model specification, and validation across a range of regression contexts*.
+<br> *Integrates statistical rigor with practical guidance on variable selection, model specification, and model validation across a range of regression contexts.
+Focuses on restricted cubic splines as tool for modeling functional forms.*.
 
@@
 [https://www.wiley.com/en-us/Multivariable+Model+-+Building%3A+A+Pragmatic+Approach+to+Regression+Anaylsis+based+on+Fractional+Polynomials+for+Modelling+Continuous+Variables-p-9780470028421](Link)
-<br> *Presents a structured approach to modeling continuous variables using fractional polynomials, with an emphasis on reproducibility and clinical relevance*.
+<br> *Presents a pragmatic, structured approach to multivariable model building, where emphasis is on modeling continuous variables using fractional polynomials.
+Focuses on interpretability and clinical relevance of statistical models.*.
 

### 2025-08-18 papers.qmd
fixed broken links

@@
 John Wiley & Sons; 2008.
-[https://www.wiley.com/en-us/Multivariable+Model+-+Building%3A+A+Pragmatic+Approach+to+Regression+Anaylsis+based+on+Fractional+Polynomials+for+Modelling+Continuous+Variables-p-9780470028421](Link)
+[Link](https://www.wiley.com/en-us/Multivariable+Model+-+Building%3A+A+Pragmatic+Approach+to+Regression+Anaylsis+based+on+Fractional+Polynomials+for+Modelling+Continuous+Variables-p-9780470028421)
 <br> *Presents a pragmatic, structured approach to multivariable model building, where emphasis is on modeling continuous variables using fractional polynomials.
@@
 
-* Boer C de. A Practical Guide to Splines revised edn. Springer: New York, 2001. [https://link.springer.com/book/9780387953663](Link)
+* Boer C de. A Practical Guide to Splines revised edn. Springer: New York, 2001. [Link](https://link.springer.com/book/9780387953663)
 <br> *Covers spline-based methods in detail, offering practical advice for incorporating flexible functional forms into regression models*.
 
-* Hastie T, Tibshirani R.. Generalized Additive Models. Chapman & Hall/CRC: New York, 1990. [https://www.routledge.com/Generalized-Additive-Models/Hastie-Tibshirani/p/book/9780412343902](Link)
+* Hastie T, Tibshirani R.. Generalized Additive Models. Chapman & Hall/CRC: New York, 1990. [Link](https://www.routledge.com/Generalized-Additive-Models/Hastie-Tibshirani/p/book/9780412343902)
 <br> *Introduces foundational concepts in additive modeling, balancing statistical theory with illustrative applied examples*.

### 2025-08-18 papers.qmd
Miscellaneous changes.

@@
 <br> *Integrates statistical rigor with practical guidance on variable selection, model specification, and model validation across a range of regression contexts.
-Focuses on restricted cubic splines as tool for modeling functional forms.*.
+Focuses on restricted cubic splines as tool for modeling functional forms*.
 
@@
 <br> *Presents a pragmatic, structured approach to multivariable model building, where emphasis is on modeling continuous variables using fractional polynomials.
-Focuses on interpretability and clinical relevance of statistical models.*.
+Focuses on interpretability and clinical relevance of statistical models*.
 

### 2025-08-19 papers.qmd
TG2 publications and further suggested publications are now separate and their own title (#)

@@
 ---
-title: "TG2 Publications"
+title: "Publications"
 ---
+
+# Topic Group 2 (TG2) Publications
+
+These publications were written as part of STRATOS-TG2. Their aim is guidance on the topic of variable and functional form selection.
 
@@
 
-# Suggested Publications
+# Further Suggested Publications
+
+Here we list a number of textbooks and papers that were written outside of STRATOS-TG2 by TG2 members or other authors.
+These materials are endorsed here because they fit the aim of guidance on selection of variables and functional forms in multivariable modeling.
 

### 2025-08-19 papers.qmd
minor edit text (250819GH)

@@
 
-These publications were written as part of STRATOS-TG2. Their aim is guidance on the topic of variable and functional form selection.
+These publications were written on behalf of STRATOS-TG2. They center around the topic group's core topic, the selection of variables and functional forms in multivariable models.
 
@@
 
-Here we list a number of textbooks and papers that were written outside of STRATOS-TG2 by TG2 members or other authors.
+Here we list a number of textbooks and papers that were not written on behalf of STRATOS-TG2.
 These materials are endorsed here because they fit the aim of guidance on selection of variables and functional forms in multivariable modeling.

### 2025-09-08 papers.qmd
added Presplines paper (GH, 250908)

@@
 These publications were written on behalf of STRATOS-TG2. They center around the topic group's core topic, the selection of variables and functional forms in multivariable models.
+
+## A Systematic Categorization of Performance Measures for Estimated Non-Linear Associations Between an Outcome and Continuous Predictors
+
+Ullmann T, Heinze G, Abrahamowicz M, Perperoglou A, Sauerbrei W, Schmid M, Dunkler D, for TG2 of the STRATOS initiative, 2025.
+A Systematic Categorization of Performance Measures for Estimated Non-Linear Associations Between an Outcome and Continuous Predictors.
+WIREs Computational Statistics 17(3), e70042.
+[https://doi.org/10.1002/wics.70042](https://doi.org/10.1002/wics.70042)
+
+<details>
+  <summary>Abstract</summary>
+
+### Abstract
+
+In regression analysis, associations between continuous predictors and the outcome are often assumed to be linear.
+However, modeling the associations as non-linear can improve model fit.
+Many flexible modeling techniques, like (fractional) polynomials and spline-based approaches, are available.
+Such methods can be systematically compared in simulation studies, which require suitable performance measures to evaluate the accuracy of the estimated curves against the true data-generating functions.
+Although various measures have been proposed in the literature, no systematic overview exists so far.
+To fill this gap, we introduce a categorization of performance measures for evaluating estimated non-linear associations between an outcome and continuous predictors.
+This categorization includes many commonly used measures.
+The measures can not only be used in simulation studies, but also in application studies to compare different estimates to each other.
+We further illustrate and compare the behavior of different performance measures through some examples and a Shiny app.
+
+</details>
 

### 2025-09-12 papers.qmd
add 1 paper (GH 250912)

@@
 
+* Lopez-Ayala, P., Riley, R.D., Collins, G.S., Zimmermann, T.
+(2025).
+Dealing with continuous variables and modelling non-linear associations in healthcare data: practical guide.
+BMJ 390, e082440.
+[https://doi.org/10.1136/bmj-2024-082440](https://doi.org/10.1136/bmj-2024-082440)
+<br> *Proper handling of continuous variables is crucial in healthcare research, for example, within regression modelling for descriptive, explanatory, or predictive purposes.
+However, inadequate methods are commonly used.
+This article highlights the importance of appropriately handling continuous variables, and illustrates the consequences of categorisation.
+This article also explains why assuming a linear relationship between the independent and dependent variable might be inappropriate, and describes how to use splines or fractional polynomials to model non-linear relationships.*.
+
 * Kipruto, E., & Sauerbrei, W.

### 2025-09-18 activities.qmd
after Roes 2025

@@
 ## Recent
+
+* Invited session at the [ROeS 2025](https://roes2025.at/index.html) (Conference of the Austro-Swiss Region of the International Biometric Society) in Graz, A, 14-18 Sep 2025.
+Topic of the session: Contrasting Bayesian and frequentist model building for descriptive research questions - a paired-design experiment of STRATOS-TG2.
+Invited speaker: [Mariana Nold](/assets/Mariana_slides_Roes_2025.pdf).
+Further speakers for TG2: [Michael Kammer](/assets/20250902_Kammer.pdf), [Aliaksandr Hubin](/assets/Aliaksandr_slides_Roes_2025.pdf), [Georg Heinze](/assets/2025-09 Roes Heinze.pdf).
+
+<details>
+  <summary>Session abstract and slides</summary>
+
+### Abstract
+
+Statistical inference is characterized by a rich epistemological diversity, underpinned by both historical and contemporary debates.
+The frequentist approach, never having been a closed theory, was characterized by conflicts within statistical methodologies.
+Similarly, significant differences and even conflicts have existed both within and between frequentist and Bayesian schools.
+Despite these tensions, efforts have been made to combine the strengths of frequentist and Bayesian approaches in a pragmatic manner, rather than adhering dogmatically to one paradigm.
+Our session aims to explore the extent to which modelling approaches are inherently frequentist or Bayesian, and whether these distinctions lead to divergent results.
+We will compare Bayesian and frequentist model building in four case studies: predicting bacteremia, estimating the proportion of body fat, describing the orbits of exoplanets, investigating the sense of school-belonging in Austrian pupils.
+For each case study, a research question suited to a multivariable regression model was defined.
+Two analysis plans were developed per study by different individuals: one Bayesian and one frequentist.
+Data is currently being analyzed in parallel, and results will be presented in the session.
+We will thoroughly discuss each step, examining the nuances of each approach to enhance understanding of the methodological and practical implications of statistical inference paradigms in complex, real-world problems.
+
+Pictures:
+
+[Starting the session](/assets/1758101359311.jpeg) (thanks to Sereina Herzog)
+
+[After the session](/assets/PXL_20250918_080447284.jpg) (thanks to Alexandra Kaider)
+
+</details>
+
+* Invited talk at the [ROeS 2025](https://roes2025.at/index.html) (Conference of the Austro-Swiss Region of the International Biometric Society) in Graz, A, 14-18 Sep 2025.
+Topic of the session: Towards best practices in medical statistics: The need to implement innovative scientific solutions.
+Speaker: Theresa Ullmann.
+Topic of the talk: Neutral Comparisons Of Statistical Methods: Opportunities, Challenges, And Common Issues (Includes as an example TG2-P11, Simulation study on variable selection)
+
+* Invited session at 70th Annual Meeting of the German Society for Medical Informatics, Biometrics and Epidemiology (Deutsche Gesellschaft für Medizinische Informatik, Biometrie und Epidemiologie e.
+V., GMDS) in Jena, Germany, 7-11 Sep 2025.
+Speaker for TG2: Gregor Buch.
+Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.
+[Find out more (in German)](https://gmds2025.de/)
+
+* Minisymposium at the International Society of Clinical Biostatistics (ISCB) conference in Basel, CH, 28 Aug 2025.
+Topic of the session: Statistical Research needs to improve – on the important roles of simulation studies and guidance for analysis.
+Speakers (among others): Georg Heinze, Willi Sauerbrei, Michal Abrahamowicz.
+[Find out more](https://iscb2025.info/mini-symposia.html#mini-symposium-3-statistical-research-needs-to-improve-on-the-important-roles-of-simulation-studies-and-guidance-for-analysis)
 
@@
 
-* Minisymposium at the International Society of Clinical Biostatistics (ISCB) conference in Basel, CH, 28 Aug 2025.
-Topic of the session: Statistical Research needs to improve – on the important roles of simulation studies and guidance for analysis.
-Speakers (among others): Georg Heinze, Willi Sauerbrei, Michal Abrahamowicz.
-[Find out more](https://iscb2025.info/mini-symposia.html#mini-symposium-3-statistical-research-needs-to-improve-on-the-important-roles-of-simulation-studies-and-guidance-for-analysis)
-
-* Invited session at 70th Annual Meeting of the German Society for Medical Informatics, Biometrics and Epidemiology (Deutsche Gesellschaft für Medizinische Informatik, Biometrie und Epidemiologie e.
-V., GMDS) in Jena, Germany, 7-11 Sep 2025.
-Speaker for TG2: Gregor Buch.
-Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.
-[Find out more (in German)](https://gmds2025.de/)
-
-* Invited session at the [ROeS 2025](https://roes2025.at/index.html) (Conference of the Austro-Swiss Region of the International Biometric Society) in Graz, A, 14-18 Sep 2025.
-Topic of the session: Contrasting Bayesian and frequentist model building for descriptive research questions - a paired-design experiment of STRATOS-TG2.
-Invited speaker: Mariana Nold, further speakers for TG2: Michael Kammer, Georg Heinze, Aliaksandr Hubin
-
-* Invited talk at the [ROeS 2025](https://roes2025.at/index.html) (Conference of the Austro-Swiss Region of the International Biometric Society) in Graz, A, 14-18 Sep 2025.
-Topic of the session: Towards best practices in medical statistics: The need to implement innovative scientific solutions.
-Speaker: Theresa Ullmann.
-Topic of the talk: Neutral Comparisons Of Statistical Methods: Opportunities, Challenges, And Common Issues (Includes as an example TG2-P11, Simulation study on variable selection)
+* summer activities completed! we are preparing for 2026...

### 2025-09-19 activities.qmd
after Roes, typo corrected (GH 250918)
add Roes slides Theresa (GH 250919)

@@
 Topic of the session: Contrasting Bayesian and frequentist model building for descriptive research questions - a paired-design experiment of STRATOS-TG2.
-Invited speaker: [Mariana Nold](/assets/Mariana_slides_Roes_2025.pdf).
-Further speakers for TG2: [Michael Kammer](/assets/20250902_Kammer.pdf), [Aliaksandr Hubin](/assets/Aliaksandr_slides_Roes_2025.pdf), [Georg Heinze](/assets/2025-09 Roes Heinze.pdf).
+Invited speaker: <a href="/assets/Mariana_slides_Roes_2025.pdf" target="_blank" rel="noopener noreferrer">Mariana Nold</a>. 
+Further speakers for TG2: <a href="/assets/20250902_Kammer.pdf" target="_blank" rel="noopener noreferrer">Michael Kammer</a>, 
+<a href="/assets/Aliaksandr_slides_Roes_2025.pdf" target="_blank" rel="noopener noreferrer">Aliaksandr Hubin</a>, 
+<a href="/assets/2025-09 Roes Heinze.pdf" target="_blank" rel="noopener noreferrer">Georg Heinze</a>.
 
 <details>
-  <summary>Session abstract and slides</summary>
+  <summary>Session abstract and pictures</summary>
 
@@
 Speaker: Theresa Ullmann.
-Topic of the talk: Neutral Comparisons Of Statistical Methods: Opportunities, Challenges, And Common Issues (Includes as an example TG2-P11, Simulation study on variable selection)
+Topic of the talk: [Neutral Comparisons Of Statistical Methods: Opportunities, Challenges, And Common Issues](assets/Slides ROeS Ullmann.pdf) (Includes as an example TG2-P11, Simulation study on variable selection)
 

### 2025-09-24 members.qmd
add FK, TP, talk of Gregor (GH 250924)
Tu moved to members (GH 250924)

@@
 
+**[Theresa Ullmann](https://www.meduniwien.ac.at/researcher/theresa_ullmann)**: Institute of Clinical Biometrics, Center for Medical Data Science, Medical University of Vienna,  Austria
+
 ## Clinical affiliate
@@
 
+**[Franziska Kappenberg](https://www.imbie.uni-bonn.de/people/dr-franziska-kappenberg/)**: Institute for Medicine Biometry, Informatics and Epidemiology, University of Bonn, Bonn, Germany
+
 **Edwin  Kipruto**:  Institute for Medical Biometry and Medical Informatics, Medical Center - University of Freiburg, Germany
 
-**[Theresa Ullmann](https://www.meduniwien.ac.at/researcher/theresa_ullmann)**: Institute of Clinical Biometrics, Center for Medical Data Science, Medical University of Vienna,  Austria
+**Thomas Prince**: Institute for Medicine Biometry, Informatics and Epidemiology, University of Bonn, Bonn, Germany

### 2025-11-25 activities.qmd
add FK, TP, talk of Gregor (GH 250924)
update (251125, GH)

@@
 ---
+
+## Upcoming
+
+* Invited talk at the [CFE-CMStatistics 2025](https://www.cmstatistics.org/CFECMStatistics2025) (19th International Joint Conference on Computational and Financial Econometrics and Computational and Methodological Statistics) in London, UK, 13-15 Dec 2025.
+Topic of the session: Researcher degrees of freedom, flexible modeling and interpretability. 
+Speaker: Theresa Ullmann. 
+Topic of the talk: Comparing methods for flexible estimation of non-linear associations: The importance of suitable performance measures (project TG2-P8).
 
@@
 Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.
-[Find out more (in German)](https://gmds2025.de/)
+[Slides](assets/Covid-Modeling-Review_Talk_GMDS2025.pdf)
 

### 2025-11-25 activities.qmd
corrected links to slides (251125 GH)

@@
 Topic of the session: Contrasting Bayesian and frequentist model building for descriptive research questions - a paired-design experiment of STRATOS-TG2.
-Invited speaker: <a href="/assets/Mariana_slides_Roes_2025.pdf" target="_blank" rel="noopener noreferrer">Mariana Nold</a>. 
-Further speakers for TG2: <a href="/assets/20250902_Kammer.pdf" target="_blank" rel="noopener noreferrer">Michael Kammer</a>, 
-<a href="/assets/Aliaksandr_slides_Roes_2025.pdf" target="_blank" rel="noopener noreferrer">Aliaksandr Hubin</a>, 
-<a href="/assets/2025-09 Roes Heinze.pdf" target="_blank" rel="noopener noreferrer">Georg Heinze</a>.
+Invited speaker: Mariana Nold <a href="/assets/Mariana_slides_Roes_2025.pdf" target="_blank" rel="noopener noreferrer">Slides</a>. 
+Further speakers for TG2: Michael Kammer <a href="/assets/20250902_Kammer.pdf" target="_blank" rel="noopener noreferrer">Slides</a>, Aliaksandr Hubin 
+<a href="/assets/Aliaksandr_slides_Roes_2025.pdf" target="_blank" rel="noopener noreferrer">Slides</a>, 
+Georg Heinze <a href="/assets/2025-09 Roes Heinze.pdf" target="_blank" rel="noopener noreferrer">Slides</a>.
 

### 2025-11-26 activities.qmd
removed 2nd upcoming (251126GH)

--- 2025-11-25 activities.qmd
@@
 * Invited talk at the [CFE-CMStatistics 2025](https://www.cmstatistics.org/CFECMStatistics2025) (19th International Joint Conference on Computational and Financial Econometrics and Computational and Methodological Statistics) in London, UK, 13-15 Dec 2025.
-Topic of the session: Researcher degrees of freedom, flexible modeling and interpretability. 
-Speaker: Theresa Ullmann. 
-Topic of the talk: Comparing methods for flexible estimation of non-linear associations: The importance of suitable performance measures (project TG2-P8).
+  * Topic of the session: Researcher degrees of freedom, flexible modeling and interpretability. 
+  * Speaker: Theresa Ullmann. 
+  * Topic of the talk: Comparing methods for flexible estimation of non-linear associations: The importance of suitable performance measures (project TG2-P8).
 
@@
 * Invited session at the [ROeS 2025](https://roes2025.at/index.html) (Conference of the Austro-Swiss Region of the International Biometric Society) in Graz, A, 14-18 Sep 2025. 
-Topic of the session: Contrasting Bayesian and frequentist model building for descriptive research questions - a paired-design experiment of STRATOS-TG2.
-Invited speaker: <a href="/assets/Mariana_slides_Roes_2025.pdf" target="_blank" rel="noopener noreferrer">Mariana Nold</a>. 
-Further speakers for TG2: <a href="/assets/20250902_Kammer.pdf" target="_blank" rel="noopener noreferrer">Michael Kammer</a>, 
-<a href="/assets/Aliaksandr_slides_Roes_2025.pdf" target="_blank" rel="noopener noreferrer">Aliaksandr Hubin</a>, 
-<a href="/assets/2025-09 Roes Heinze.pdf" target="_blank" rel="noopener noreferrer">Georg Heinze</a>.
+  * Topic of the session: Contrasting Bayesian and frequentist model building for descriptive research questions - a paired-design experiment of STRATOS-TG2. 
+  * Invited speaker: Mariana Nold <a href="/assets/Mariana_slides_Roes_2025.pdf" target="_blank" rel="noopener noreferrer">Slides</a>. 
+  * Further speakers for TG2: 
+    * Michael Kammer <a href="/assets/20250902_Kammer.pdf" target="_blank" rel="noopener noreferrer">Slides</a>, 
+    * Aliaksandr Hubin 
+<a href="/assets/Aliaksandr_slides_Roes_2025.pdf" target="_blank" rel="noopener noreferrer">Slides</a>, 
+    * Georg Heinze <a href="/assets/2025-09 Roes Heinze.pdf" target="_blank" rel="noopener noreferrer">Slides</a>.
 
@@
 * Invited talk at the [ROeS 2025](https://roes2025.at/index.html) (Conference of the Austro-Swiss Region of the International Biometric Society) in Graz, A, 14-18 Sep 2025. 
-Topic of the session: Towards best practices in medical statistics: The need to implement innovative scientific solutions.
-Speaker: Theresa Ullmann.
-Topic of the talk: [Neutral Comparisons Of Statistical Methods: Opportunities, Challenges, And Common Issues](assets/Slides ROeS Ullmann.pdf) (Includes as an example TG2-P11, Simulation study on variable selection)
+  * Topic of the session: Towards best practices in medical statistics: The need to implement innovative scientific solutions. 
+  * Speaker: Theresa Ullmann. 
+  * Topic of the talk: [Neutral Comparisons Of Statistical Methods: Opportunities, Challenges, And Common Issues](assets/Slides ROeS Ullmann.pdf) (Includes as an example TG2-P11, Simulation study on variable selection)
 
@@
 V., GMDS) in Jena, Germany, 7-11 Sep 2025.
-Speaker for TG2: Gregor Buch.
-Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.
+  * Speaker for TG2: Gregor Buch. 
+  * Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.
 [Slides](assets/Covid-Modeling-Review_Talk_GMDS2025.pdf)
@@
 * Minisymposium at the International Society of Clinical Biostatistics (ISCB) conference in Basel, CH, 28 Aug 2025. 
-Topic of the session: Statistical Research needs to improve – on the important roles of simulation studies and guidance for analysis.
-Speakers (among others): Georg Heinze, Willi Sauerbrei, Michal Abrahamowicz.
+  * Topic of the session: Statistical Research needs to improve – on the important roles of simulation studies and guidance for analysis. 
+  * Speakers (among others): Georg Heinze, Willi Sauerbrei, Michal Abrahamowicz.
 [Find out more](https://iscb2025.info/mini-symposia.html#mini-symposium-3-statistical-research-needs-to-improve-on-the-important-roles-of-simulation-studies-and-guidance-for-analysis)
 
-* Topic-contributed paper session at Joint Statistical Meeting (JSM), 6 Aug 2025.
-Speaker for TG2: Marc Henrion.
-Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.[Find out more](https://ww2.amstat.org/meetings/jsm/2025/)
+* Topic-contributed paper session at Joint Statistical Meeting (JSM), 6 Aug 2025. Speaker for TG2: Marc Henrion. 
+  * Topic of the talk: STRATOS Topic Group 2 – Selection of variables and functional forms in multivariable analysis: Results from a review of Covid-19 prognostic models.[Find out more](https://ww2.amstat.org/meetings/jsm/2025/)
 
@@
 [The slides are available at OSF here](https://osf.io/a5gtk/).
-
-## Upcoming
-
-* summer activities completed! we are preparing for 2026...
